---
title: "StarDist Segmentation QC"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
.libPaths("/mnt/project/blfridley/PCF_pilots/_common/R/renv/library/linux-rhel-8.10/R-4.4/x86_64-pc-linux-gnu")

suppressPackageStartupMessages({
  library(yaml)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(tidyr)
  library(kableExtra)
  library(stringr)
  library(tibble)
  library(viridis)
  library(patchwork)
})

knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  comment = NA,
  fig.width  = 12,
  fig.height = 6
)

cfg_path <- "config.yml"
cfg      <- yaml::read_yaml(cfg_path)

FLAG_RED    <- "#d62728"
FLAG_ORANGE <- "#ff7f0e"
NEUTRAL     <- "#aec7e8"
```

```{r load-data, cache=FALSE}
data_processed <- fread(
  file.path(dirname(cfg_path), cfg$paths$folders$results, "data_processed.csv")
) %>% as.data.frame()

nzNormedMarkers <- grep("^nzNorm_", names(data_processed), value = TRUE)
markers         <- sub("^nzNorm_", "", nzNormedMarkers)

qc_cols <- names(data_processed)[str_detect(
  names(data_processed),
  regex("Area|Perimeter|Circularity|caliper|Eccentricity|ratio|Length|diameter|Solidity",
        ignore_case = TRUE)
)]

x <- data_processed %>%
  select(SampleID, ImageID, all_of(qc_cols)) %>%
  filter(if_all(all_of(qc_cols), ~ !is.na(.))) %>%
  mutate(across(all_of(qc_cols), as.numeric)) %>%
  mutate(across(
    matches("Area|Perimeter|Length|diameter|caliper", ignore.case = TRUE),
    ~ log10(.x + 1)
  ))

core_feats <- c(
  "Nucleus: Area µm^2",
  "Nucleus: Length µm",
  "Nucleus: Circularity",
  "Nucleus: Solidity",
  "Nucleus: Max diameter µm",
  "Nucleus: Min diameter µm",
  "Cell: Area µm^2",
  "Cell: Length µm",
  "Cell: Circularity",
  "Cell: Solidity",
  "Cell: Max diameter µm",
  "Cell: Min diameter µm",
  "Nucleus/Cell area ratio"
)

x <- x %>%
  mutate(across(all_of(intersect(core_feats, names(x))), as.numeric))
```

# Overview

This report evaluates the quality of StarDist-based cell segmentation across
all samples prior to marker expression analysis. The goal is to identify
samples or images with anomalous morphology distributions, extreme measurement
tails, or geometric inconsistencies that could compromise downstream phenotyping.

Sections are ordered from population-level to sample-level to cell-level:
cell counts first, then morphology distributions, then bivariate sanity
checks, then the flagging table. Marker expression densities are included
at the end to inform GammaGateR boundary specification.

**What requires action after reviewing this report:**

- Samples flagged in the QC table should be inspected in QuPath before
  proceeding to GammaGateR fitting.
- Marker density plots should be used to set mode constraint bounds in
  `run_GGR.R`.

```{r overview-table}
tibble(
  Metric  = c("Total samples", "Total images", "Total cells", "Markers"),
  Value   = c(
    as.character(n_distinct(x$SampleID)),
    as.character(n_distinct(x$ImageID)),
    format(nrow(x), big.mark = ","),
    as.character(length(markers))
  )
) %>%
  kbl(align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, position = "left") %>%
  column_spec(1, bold = TRUE)
```

---

# 1. Cell Counts

Cell counts per image and per sample provide the first indication of
segmentation quality. Images with unusually low counts may have failed
segmentation, poor tissue quality, or overly restrictive detection
parameters. Images with unusually high counts may have segmentation
over-splitting artifacts.

**What good looks like:** Counts are broadly consistent across images
within the same sample. Large between-sample differences are expected
if tissue areas differ.

**Red flags:** Any image with near-zero cells, or any sample where one
image has drastically fewer cells than others from the same tissue block.

```{r cell-counts-image, fig.height=7}
x %>%
  count(ImageID, SampleID, name = "n_cells") %>%
  ggplot(aes(x = reorder(ImageID, n_cells), y = n_cells, fill = SampleID)) +
  geom_col(color = "white", linewidth = 0.2) +
  geom_text(aes(label = format(n_cells, big.mark = ",")),
            hjust = -0.1, size = 3) +
  coord_flip(clip = "off") +
  expand_limits(y = max(x %>% count(ImageID) %>% pull(n)) * 1.15) +
  scale_fill_viridis_d(option = "turbo", name = "Sample") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(title = "Detected cells per image",
       x = NULL, y = "Cell count") +
  theme_bw(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

```{r cell-counts-sample, fig.height=4}
x %>%
  count(SampleID, name = "n_cells") %>%
  ggplot(aes(x = reorder(SampleID, n_cells), y = n_cells)) +
  geom_col(fill = NEUTRAL, color = "grey40", linewidth = 0.3) +
  geom_text(aes(label = format(n_cells, big.mark = ",")),
            hjust = -0.1, size = 3) +
  coord_flip(clip = "off") +
  expand_limits(y = max(x %>% count(SampleID) %>% pull(n)) * 1.15) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(title = "Total detected cells per sample",
       x = NULL, y = "Cell count") +
  theme_bw(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

---

# 2. Morphology Feature Distributions

## 2a. Univariate distributions

Histograms and density curves for core morphology features pooled across
all samples. Area and diameter features are log10-transformed prior to
plotting. Well-segmented data typically shows unimodal, roughly symmetric
distributions for area and diameter features, and left-skewed distributions
for circularity and solidity (most cells near 1).

**Red flags:** Bimodal distributions in area or diameter suggest two
distinct cell populations or a segmentation artifact splitting large cells
into fragments. Heavy left tails in circularity or solidity indicate
irregular or fragmented segmentation objects.

```{r univariate-distributions, fig.height=8}
x %>%
  select(SampleID, all_of(intersect(core_feats, names(x)))) %>%
  pivot_longer(-SampleID, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = NEUTRAL, color = "grey40",
                 linewidth = 0.2) +
  geom_density(linewidth = 0.6, color = "#1f3a5f") +
  facet_wrap(~ feature, scales = "free", ncol = 4) +
  labs(title = "Distributions of core morphology features",
       subtitle = "Area and diameter features are log10-transformed",
       x = NULL, y = "Density") +
  theme_bw(base_size = 10) +
  theme(strip.text   = element_text(face = "bold", size = 8),
        plot.title   = element_text(face = "bold"),
        panel.spacing = unit(8, "pt"))
```

## 2b. Per-sample boxplots

Sample-level distributions of key morphology features. Consistent medians
and spreads across samples indicate stable segmentation. Outlier samples
should be flagged for inspection.

**Red flags:** Any sample with a substantially shifted median or compressed
interquartile range relative to others for the same feature.

```{r per-sample-boxplots, fig.height=8}
x %>%
  select(SampleID,
         all_of(intersect(c(
           "Nucleus: Area µm^2", "Nucleus: Circularity",
           "Nucleus: Solidity", "Cell: Area µm^2",
           "Cell: Circularity", "Cell: Solidity",
           "Nucleus/Cell area ratio"
         ), names(x)))) %>%
  pivot_longer(-SampleID, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = SampleID, y = value)) +
  geom_boxplot(fill = NEUTRAL, outlier.size = 0.3,
               outlier.alpha = 0.3, linewidth = 0.4) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  labs(title = "Morphology features by sample",
       x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
        strip.text   = element_text(face = "bold", size = 8),
        plot.title   = element_text(face = "bold"),
        panel.spacing = unit(8, "pt"))
```

---

# 3. Bivariate Sanity Checks

## 3a. Area vs Circularity

Cell area and circularity should be weakly negatively correlated — larger
cells tend to be slightly less circular. A strong negative correlation or
a discrete cluster of low-circularity large objects suggests segmentation
of non-cellular artifacts (tissue debris, folded tissue).

```{r area-vs-circularity, fig.height=5}
x %>%
  select(SampleID,
         `Cell: Area µm^2`, `Cell: Circularity`,
         `Nucleus: Area µm^2`, `Nucleus: Circularity`) %>%
  pivot_longer(
    cols = -SampleID,
    names_to  = c("Compartment", ".value"),
    names_pattern = "(Cell|Nucleus): (.*)"
  ) %>%
  ggplot(aes(`Area µm^2`, Circularity, color = SampleID)) +
  geom_point(alpha = 0.2, size = 0.4, shape = 16) +
  scale_x_continuous() +
  scale_color_viridis_d(option = "turbo", name = "Sample") +
  facet_wrap(~ Compartment, scales = "free_x") +
  labs(title = "Area vs Circularity by compartment",
       x = "Area µm² (log10)", y = "Circularity") +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"))
```

## 3b. Feature correlation heatmap

Spearman correlation between all segmentation features. Highly correlated
features (|r| > 0.9) are redundant and dominated by the same underlying
measurement. Unexpected correlations between nucleus and cell features may
indicate segmentation boundary issues.

```{r correlation-heatmap, fig.height=7}
x %>%
  select(all_of(qc_cols)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman") %>%
  as.data.frame() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "r") %>%
  ggplot(aes(Var1, Var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low  = "#2166ac", mid = "white", high = "#d62728",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Spearman r") +
  coord_equal() +
  labs(title = "Spearman correlation of segmentation morphology features",
       x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y  = element_text(size = 8),
        panel.grid   = element_blank(),
        plot.title   = element_text(face = "bold"))
```

## 3c. PCA

PCA of all segmentation features colored by SampleID. Samples that cluster
separately from all others may have systematically different segmentation
behavior — larger nuclei, different staining intensity calibration, or
tissue-specific morphology. This is not automatically a problem but warrants
inspection if a sample is a clear outlier.

```{r pca, fig.height=5}
pcs <- prcomp(
  x %>% select(all_of(qc_cols)),
  center = TRUE, scale. = TRUE
)

pct_var <- round(pcs$sdev^2 / sum(pcs$sdev^2) * 100, 1)

bind_cols(x, as.data.frame(pcs$x[, 1:2])) %>%
  ggplot(aes(PC1, PC2, color = SampleID)) +
  geom_point(alpha = 0.35, size = 0.5, shape = 16) +
  scale_color_viridis_d(option = "turbo", name = "Sample") +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(
    title = "PCA of segmentation morphology features",
    x     = paste0("PC1 (", pct_var[1], "% variance)"),
    y     = paste0("PC2 (", pct_var[2], "% variance)")
  ) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

# 4. Segmentation QC Flag Table

Per-sample summary of extreme measurements. Each metric represents the
fraction of cells in that sample exceeding the stated threshold. Thresholds
are chosen to catch genuine segmentation pathologies rather than biological
outliers.

- **Extreme area tails** — fraction of cells below the 0.5th or above the
  99.5th percentile of nuclear or cell area. Values above 5% suggest the
  distribution has heavy tails from debris or over-segmentation.
- **Low circularity** — highly non-circular objects are likely tissue folds,
  debris, or segmentation artifacts joining adjacent cells.
- **Low solidity** — concave objects suggest irregular boundaries from
  poor segmentation or non-cellular material.
- **High nucleus/cell ratio** — nucleus area exceeding 95% of cell area is
  geometrically implausible and indicates a segmentation boundary error.
- **Invalid ratio** — nucleus/cell ratio ≤ 0 or > 1 is mathematically
  impossible and indicates a data export or segmentation error.

```{r qc-flag-table}
qc <- x %>%
  select(SampleID,
         all_of(intersect(c(
           "Nucleus: Area µm^2", "Cell: Area µm^2",
           "Nucleus: Circularity", "Cell: Circularity",
           "Nucleus: Solidity", "Cell: Solidity",
           "Nucleus/Cell area ratio"
         ), names(x)))) %>%
  group_by(SampleID) %>%
  summarise(
    n_cells = n(),
    p_nuc_area_tail = mean(
      `Nucleus: Area µm^2` < quantile(`Nucleus: Area µm^2`, 0.005, na.rm = TRUE) |
      `Nucleus: Area µm^2` > quantile(`Nucleus: Area µm^2`, 0.995, na.rm = TRUE),
      na.rm = TRUE),
    p_cell_area_tail = mean(
      `Cell: Area µm^2` < quantile(`Cell: Area µm^2`, 0.005, na.rm = TRUE) |
      `Cell: Area µm^2` > quantile(`Cell: Area µm^2`, 0.995, na.rm = TRUE),
      na.rm = TRUE),
    p_nuc_circ_low   = mean(`Nucleus: Circularity`    < 0.20, na.rm = TRUE),
    p_cell_circ_low  = mean(`Cell: Circularity`       < 0.15, na.rm = TRUE),
    p_nuc_solid_low  = mean(`Nucleus: Solidity`       < 0.80, na.rm = TRUE),
    p_cell_solid_low = mean(`Cell: Solidity`          < 0.80, na.rm = TRUE),
    p_ratio_high     = mean(`Nucleus/Cell area ratio` > 0.95, na.rm = TRUE),
    p_ratio_invalid  = mean(
      `Nucleus/Cell area ratio` <= 0 | `Nucleus/Cell area ratio` > 1,
      na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(flag_any =
    p_nuc_area_tail  > 0.05 |
    p_cell_area_tail > 0.05 |
    p_ratio_high     > 0.05 |
    p_ratio_invalid  > 0.001
  )

qc %>%
  rename(
    `N cells`                            = n_cells,
    `Nucleus area tail (%)`              = p_nuc_area_tail,
    `Cell area tail (%)`                 = p_cell_area_tail,
    `Low nucleus circularity (%)`        = p_nuc_circ_low,
    `Low cell circularity (%)`           = p_cell_circ_low,
    `Low nucleus solidity (%)`           = p_nuc_solid_low,
    `Low cell solidity (%)`              = p_cell_solid_low,
    `High nucleus/cell ratio (%)`        = p_ratio_high,
    `Invalid nucleus/cell ratio (%)`     = p_ratio_invalid,
    `Any flag`                           = flag_any
  ) %>%
  mutate(across(where(is.numeric) & !`N cells`, ~ round(.x * 100, 2))) %>%
  arrange(desc(`Any flag`), SampleID) %>%
  kbl(caption = "Sample-level segmentation QC summary. Numeric columns show % of cells exceeding threshold.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(
    which(names(.) == "Any flag"),
    bold  = TRUE,
    color = if_else(qc$flag_any, FLAG_RED, OK_GREEN <- "#2ca02c")
  ) %>%
  scroll_box(width = "100%")
```

---

# 5. Normalized Expression Densities

Density curves of log-mean normalized expression per marker, one curve per
sample. These plots are used to set the mode constraint bounds for the
GammaGateR mixture model in `run_GGR.R`.

**What good looks like:** A clear bimodal or right-skewed distribution with
a dominant negative peak near 0–0.5 and a secondary positive shoulder or
peak at higher intensities.

**Red flags:** A purely unimodal distribution with no right shoulder means
the marker may be unexpressed or the normalization has compressed the dynamic
range. Large between-sample shifts in the negative peak location suggest
batch effects in staining intensity that normalization did not fully correct.

```{r marker-densities, results='asis'}
long_data <- data_processed %>%
  select(SampleID, all_of(nzNormedMarkers)) %>%
  pivot_longer(cols = all_of(nzNormedMarkers),
               names_to = "Marker", values_to = "Intensity")

cat("## {.tabset}\n\n")

for (mk in nzNormedMarkers) {
  mk_lab <- sub("^nzNorm_", "", mk)
  cat("### ", mk_lab, "\n\n", sep = "")

  print(
    long_data %>%
      filter(Marker == mk) %>%
      ggplot(aes(x = Intensity, color = SampleID)) +
      geom_density(linewidth = 0.6, alpha = 0.8) +
      scale_color_viridis_d(option = "turbo", guide = "none") +
      geom_vline(xintercept = 0.5, linetype = "dashed",
                 color = FLAG_RED, linewidth = 0.4) +
      labs(
        title    = mk_lab,
        subtitle = "Dashed line at 0.5 for reference only",
        x        = "Normalized intensity",
        y        = "Density"
      ) +
      theme_bw(base_size = 12) +
      theme(plot.title = element_text(face = "bold"))
  )
  cat("\n\n")
}
```

---

# Session Information

```{r session}
sessionInfo()
```
