---
title: "Multiplex Imaging — Binarization, Phenotyping, and QC"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    self_contained: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}

.libPaths("/mnt/project/blfridley/PCF_pilots/_common/R/renv/library/linux-rhel-8.10/R-4.4/x86_64-pc-linux-gnu")

suppressPackageStartupMessages({
  library(yaml)
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(gtsummary)
  library(purrr)
  library(tibble)
  library(scales)
  library(skimr)
  library(forcats)
  library(rlang)
  library(tools)
  library(kableExtra)
  library(ggrepel)
  library(patchwork)
  library(rstatix)
  library(stringr)
  library(parallel)
})

knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = NA
)

cfg_path <- normalizePath("../config.yml", mustWork = TRUE)
cfg      <- read_yaml(cfg_path)

FLAG_RED    <- "#d62728"
FLAG_ORANGE <- "#ff7f0e"
NEUTRAL     <- "#aec7e8"

sample_col    <- "SampleID"
lineage_col   <- "Lineage"
phenotype_col <- "Phenotype"
```

# Overview

This document performs three sequential steps following GammaGateR model
fitting:

1. **Marker binarization** — converts continuous per-cell posterior
   probabilities into binary 0/1 calls using a Bayesian FDR thresholding
   approach, with per-marker parameter control.
2. **Phenotype assignment** — applies a hierarchical rule set to the binary
   marker calls to assign each cell a lineage and phenotype label.
3. **QC and abundance summarization** — audits the gating results, flags
   anomalies, and constructs the per-sample abundance data frame for
   downstream analysis.

**Before running this document** you should have reviewed the GammaGateR
diagnostics report (`03-GGR_Diagnostics.Rmd`) and confirmed that the model
fit is adequate for each marker × sample combination.

**Sections requiring manual specification:**

- **Section 1b** — per-marker FDR targets, minimum positive rates, and
  maximum positive rate caps.
- **Section 2b** — phenotype rule hierarchy and phenotype level ordering.

Everything else runs automatically once those inputs are set.

---

# 1. Marker Binarization

## 1a. Load posterior probabilities

Each row in `df_base` is one cell. Columns are per-marker posterior
probabilities P(marker-positive | observed intensity) from GammaGateR,
plus cell-level metadata (ImageID, SampleID, SubjectID, X, Y).

```{r load-data}
model_version <- "fit13"

df_base <- {
  fit_dir <- normalizePath(
    file.path("..", "results", "GammaGateR", model_version),
    mustWork = TRUE
  )

  left_cols <- c("ImageID", "SubjectID", "X", "Y")

  data_processed <- fread(
    file.path("..", cfg$paths$folders$results, "data_processed.csv")
  ) %>% as_tibble()

  rds_files <- list.files(fit_dir, pattern = "\\.rds$", full.names = TRUE)
  rds_files <- rds_files[!grepl("boundar", basename(rds_files),
                                ignore.case = TRUE)]

  if (!length(rds_files))
    stop("No GammaGateR RDS files found in: ", fit_dir)

  exprZ_all <- map_dfr(rds_files, function(f) {
    fit <- readRDS(f)
    if (is.list(fit) && length(fit) == 1L) fit <- fit[[1L]]
    if (is.null(fit$expressionZ))
      stop("Missing `expressionZ` in: ", basename(f))
    as.data.frame(fit$expressionZ) %>%
      as_tibble() %>%
      mutate(SampleID  = file_path_sans_ext(basename(f)),
             CellIndex = row_number())
  })

  metadata_by_cell <- data_processed %>%
    mutate(
      SampleID_join = as.character(SampleID),
      CellIndex     = ave(seq_len(n()), SampleID_join, FUN = seq_along)
    ) %>%
    select(SampleID_join, CellIndex, any_of(left_cols))

  out <- exprZ_all %>%
    left_join(metadata_by_cell,
              by = c("SampleID" = "SampleID_join",
                     "CellIndex" = "CellIndex"))

  n_missing <- sum(is.na(out$ImageID) | is.na(out$SubjectID))
  if (n_missing > 0)
    stop("Metadata join failed for ", n_missing, " rows. ",
         "RDS cell order does not align with data_processed within SampleID.")

  out %>%
    select(-CellIndex) %>%
    rename_with(~ gsub("^nzNorm_", "", .x), starts_with("nzNorm_")) %>%
    relocate(any_of(c("ImageID", "SampleID", "SubjectID", "X", "Y")),
             .before = 1)
}

cat("df_base:", format(nrow(df_base), big.mark = ","), "cells x",
    ncol(df_base), "columns\n")
cat("Samples:", paste(sort(unique(df_base$SampleID)), collapse = ", "), "\n")
```

## 1b. Gating parameters

**Specify per-marker values here.** These three parameters jointly control
the Bayesian FDR thresholding behavior for each marker:

| Parameter | Role |
|---|---|
| `marker_fdr` | Maximum tolerated Bayesian FDR among positive calls. Lower = more stringent. Typical range 0.05–0.20. |
| `marker_min_pos` | Floor on positive call rate. Prevents rare but real populations from being thresholded to zero. |
| `marker_max_cap` | Ceiling on positive call rate. Prevents non-specific absorption of bulk negative cells. |

Consult the GammaGateR diagnostics report (λ estimates, posterior
distributions, and confidence indices) to inform these values before running.

```{r gating-params}
# ── USER-SPECIFIED ────────────────────────────────────────────────────────────

marker_fdr <- c(
  CD3e     = 0.10, CD4   = 0.20, CD8   = 0.20, FOXP3 = 0.20, CD39  = 0.20,
  CD56     = 0.05, CD20  = 0.15,
  CD66     = 0.10, MPO   = 0.10,
  CD68     = 0.05, CD163 = 0.10, CD206 = 0.10, iNOS  = 0.10,
  CD11b    = 0.10, CD14  = 0.10, CD11c = 0.10,
  `HLA-DR` = 0.10
)

marker_min_pos <- setNames(rep(0.001, length(marker_fdr)), names(marker_fdr))

marker_max_cap <- c(
  CD3e     = 0.05, CD4   = 0.05, CD8   = 0.05, FOXP3 = 0.05, CD39  = 0.05,
  CD56     = 0.05, CD20  = 0.08,
  CD66     = 0.15, MPO   = 0.15,
  CD68     = 0.20, CD163 = 0.25, CD206 = 0.25, iNOS  = 0.20,
  CD11b    = 0.40, CD14  = 0.30, CD11c = 0.20,
  `HLA-DR` = 0.50
)

```

```{r gating-params-table, echo=FALSE}
tibble(
  Marker         = names(marker_fdr),
  `FDR target`   = marker_fdr,
  `Min pos rate` = marker_min_pos,
  `Max pos cap`  = marker_max_cap
) %>%
  kbl(caption = "Gating parameters as currently specified") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  scroll_box(width = "100%")
```

## 1c. Binarization function

`make_marker_binary()` converts posterior probabilities to binary 0/1 calls.
This pipeline uses `fdr_bayesian` exclusively. The Bayesian FDR at threshold
_t_ is:

> BFDR(t) = Σ\_{i: p_i ≥ t} (1 − p_i) / |{i: p_i ≥ t}|

The largest _t_ satisfying BFDR(t) ≤ `fdr` and positive call rate within
`[min_pos_rate, max_pos_rate_cap]` is selected per marker × sample. If no
such _t_ exists, the function falls back to the `1 − min_pos_rate` quantile
and records the event for QC inspection.

```{r make-marker-binary-fn, include=FALSE}
make_marker_binary <- function(df,
                               markers,
                               threshold           = 0.5,
                               group_col           = "SampleID",
                               method              = c("fixed", "quantile",
                                                       "fdr", "fdr_bayesian"),
                               max_pos_rate        = NULL,
                               fdr                 = 0.01,
                               neg_frac            = 0.90,
                               min_pos_rate        = 5e-4,
                               max_pos_rate_cap    = 0.50,
                               return_diagnostics  = FALSE,
                               diagnostics_markers = NULL,
                               diagnostics_groups  = NULL,
                               n_grid              = 500L,
                               n_workers           = min(parallel::detectCores() - 1L, 8L),
                               eps                 = 1e-6) {

  method      <- match.arg(method)
  hit_markers <- markers[markers %in% names(df)]
  if (!length(hit_markers)) stop("No marker columns found.")
  if (!group_col %in% names(df)) stop("group_col not found in df.")

  .get_marker_val <- function(x, m, default = NULL) {
    if (is.null(x)) return(default)
    if (length(x) == 1L && is.null(names(x))) return(as.numeric(x))
    if (!is.null(names(x)) && m %in% names(x)) return(as.numeric(x[[m]]))
    default
  }

  gsym <- sym(group_col)

  .clamp_thr <- function(thr, v) {
    if (!is.finite(thr)) return(thr)
    vmax <- max(v, na.rm = TRUE)
    if (is.finite(vmax)) thr <- min(thr, vmax - eps)
    min(thr, 1 - eps)
  }

  .enforce_mpr <- function(thr0, v, mpr) {
    if (!is.finite(thr0)) return(thr0)
    mpr     <- max(min(as.numeric(mpr), 0.99), 1e-12)
    thr_mpr <- as.numeric(quantile(v, probs = 1 - mpr,
                                   na.rm = TRUE, names = FALSE, type = 1))
    min(thr0, thr_mpr)
  }

  # ── Fixed threshold ──────────────────────────────────────────────────────────
  if (method == "fixed") {
    thr <- if (length(threshold) == 1L) {
      setNames(rep(threshold, length(hit_markers)), hit_markers)
    } else if (!is.null(names(threshold))) {
      out <- setNames(rep(0.5, length(hit_markers)), hit_markers)
      out[intersect(names(threshold), hit_markers)] <-
        threshold[intersect(names(threshold), hit_markers)]
      out
    } else {
      if (length(threshold) != length(hit_markers))
        stop("Unnamed `threshold` must match length of found markers.")
      setNames(threshold, hit_markers)
    }

    out_df <- df %>%
      select(any_of(c("ImageID", "SampleID", "SubjectID", "X", "Y")),
             all_of(hit_markers)) %>%
      mutate(across(all_of(hit_markers),
                    ~ as.integer(.x >= thr[cur_column()])))

    if (!return_diagnostics) return(out_df)
    return(list(
      out         = out_df,
      diagnostics = list(method = method, group_col = group_col,
                         hit_markers = hit_markers,
                         thr_tbl = tibble(), thr_long = tibble())
    ))
  }

  # ── Quantile threshold ───────────────────────────────────────────────────────
  f_quant_thr <- function(v, r, mpr, cap) {
    v <- v[is.finite(v)]
    if (!length(v)) return(NA_real_)
    r    <- max(min(as.numeric(r), as.numeric(cap)), as.numeric(mpr))
    thr0 <- as.numeric(quantile(v, probs = 1 - r,
                                na.rm = TRUE, names = FALSE, type = 7))
    .clamp_thr(thr0, v)
  }

  # ── Frequentist FDR threshold ────────────────────────────────────────────────
  f_fdr_thr <- function(v, alpha, nf, mpr, cap) {
    alpha <- max(min(as.numeric(alpha), 0.5),  1e-6)
    nf    <- max(min(as.numeric(nf),    0.99), 0.50)
    mpr   <- max(min(as.numeric(mpr),   0.99), 1e-6)
    cap   <- max(min(as.numeric(cap),   0.99), mpr)
    v     <- v[is.finite(v)]
    if (!length(v)) return(NA_real_)

    sv <- sort(v); n <- length(sv)
    ts <- rle(sv)$values
    if (length(ts) > n_grid)
      ts <- sort(unique(as.numeric(quantile(
        ts, probs = seq(0.001, 0.999, length.out = n_grid),
        na.rm = TRUE, names = FALSE, type = 7
      ))))
    tmax <- max(ts, na.rm = TRUE); ts <- ts[ts < tmax]

    qneg   <- as.numeric(quantile(v, probs = nf,
                                  na.rm = TRUE, names = FALSE, type = 7))
    sv_neg <- sv[sv <= qneg]

    if (!length(ts) || !length(sv_neg)) {
      thr0 <- as.numeric(quantile(v, probs = 1 - mpr,
                                  na.rm = TRUE, names = FALSE, type = 7))
      return(.clamp_thr(.enforce_mpr(thr0, v, mpr), v))
    }

    rr    <- rle(sv);  u <- rr$values; cnt <- rr$lengths; cum <- cumsum(cnt)
    rrn   <- rle(sv_neg); u_n <- rrn$values; cnt_n <- rrn$lengths
    cum_n <- cumsum(cnt_n); n_neg <- length(sv_neg)

    le_idx   <- findInterval(ts, u)
    le_cnt   <- ifelse(le_idx > 0L, cum[le_idx], 0L)
    mtch     <- match(ts, u, nomatch = 0L)
    eq_cnt   <- ifelse(mtch > 0L, cnt[mtch], 0L)
    ge_cnt   <- n - (le_cnt - eq_cnt); den <- ge_cnt / n

    le_idx_n <- findInterval(ts, u_n)
    le_cnt_n <- ifelse(le_idx_n > 0L, cum_n[le_idx_n], 0L)
    mtch_n   <- match(ts, u_n, nomatch = 0L)
    eq_cnt_n <- ifelse(mtch_n > 0L, cnt_n[mtch_n], 0L)
    ge_cnt_n <- n_neg - (le_cnt_n - eq_cnt_n); num <- ge_cnt_n / n_neg

    fdr_hat <- ifelse(den > 0, num / den, 1)
    ok      <- which(fdr_hat <= alpha & den >= mpr & den <= cap)
    thr0    <- if (!length(ok))
      as.numeric(quantile(v, probs = 1 - mpr,
                          na.rm = TRUE, names = FALSE, type = 7))
    else ts[max(ok)]
    .clamp_thr(.enforce_mpr(thr0, v, mpr), v)
  }

  # ── Bayesian FDR threshold ───────────────────────────────────────────────────
  # BFDR(t) = sum_{i: p_i >= t}(1 - p_i) / |{i: p_i >= t}|
  # Optimised: findInterval replaces vapply loop (O(n log n) vs O(n * |ts|)).
  # Threshold grid capped at n_grid points; cum_false precomputed once.
  f_bfdr_thr <- function(v, alpha, mpr, cap) {
    alpha <- max(min(as.numeric(alpha), 0.5),  1e-6)
    mpr   <- max(min(as.numeric(mpr),   0.99), 1e-6)
    cap   <- max(min(as.numeric(cap),   0.99), mpr)
    v     <- sort(v[is.finite(v)])
    n     <- length(v)
    if (!n) return(NA_real_)

    ts <- unique(v)
    if (length(ts) > n_grid)
      ts <- unique(as.numeric(quantile(
        ts, probs = seq(0.001, 0.999, length.out = n_grid),
        na.rm = TRUE, names = FALSE, type = 7
      )))
    ts <- sort(ts[ts < max(ts)])

    if (!length(ts)) {
      thr0 <- as.numeric(quantile(v, probs = 1 - mpr,
                                  na.rm = TRUE, names = FALSE, type = 7))
      return(.clamp_thr(.enforce_mpr(thr0, v, mpr), v))
    }

    # k_of_t[i] = number of cells with value >= ts[i]
    # v is sorted ascending; findInterval gives last index < ts[i]
    k_of_t    <- n - findInterval(ts, v, left.open = TRUE)
    den_t     <- k_of_t / n

    # Precompute cumulative sum of (1 - p) over descending-sorted values
    # cum_false[k] = expected false positives among top-k cells
    cum_false <- cumsum(1 - rev(v))
    bfdr_t    <- ifelse(k_of_t > 0L, cum_false[k_of_t] / k_of_t, 1)

    ok   <- which(bfdr_t <= alpha & den_t >= mpr & den_t <= cap)
    thr0 <- if (!length(ok))
      as.numeric(quantile(v, probs = 1 - mpr,
                          na.rm = TRUE, names = FALSE, type = 7))
    else ts[max(ok)]
    .clamp_thr(.enforce_mpr(thr0, v, mpr), v)
  }

  # ── Compute thresholds ───────────────────────────────────────────────────────
  # Thresholds computed directly from wide df without pivot_longer,
  # avoiding materialising an n_cells * n_markers long-format frame.
  # Groups run in parallel via furrr when n_workers > 1.

  sample_ids <- unique(df[[group_col]])

  compute_group_thr <- function(m, s) {
    v     <- df[[m]][df[[group_col]] == s]
    v     <- v[is.finite(v)]
    alpha <- as.numeric(.get_marker_val(fdr,             m, fdr))
    nf    <- as.numeric(.get_marker_val(neg_frac,         m, neg_frac))
    mpr   <- as.numeric(.get_marker_val(min_pos_rate,     m, min_pos_rate))
    cap   <- as.numeric(.get_marker_val(max_pos_rate_cap, m, max_pos_rate_cap))

    thr_val <- switch(method,
      quantile = {
        r <- as.numeric(.get_marker_val(max_pos_rate, m, max_pos_rate))
        f_quant_thr(v, r = r, mpr = mpr, cap = cap)
      },
      fdr          = f_fdr_thr(v,  alpha = alpha, nf = nf, mpr = mpr, cap = cap),
      fdr_bayesian = f_bfdr_thr(v, alpha = alpha, mpr = mpr, cap = cap)
    )

    tibble(
      !!group_col := s,
      marker      = m,
      n           = length(v),
      alpha       = alpha,
      nf          = if (method == "fdr") nf else NA_real_,
      mpr         = mpr,
      cap         = cap,
      thr         = thr_val,
      q_neg       = if (method == "fdr")
        as.numeric(quantile(v, probs = nf, na.rm = TRUE,
                            names = FALSE, type = 7))
      else NA_real_
    )
  }

  groups <- expand.grid(marker = hit_markers, sample = sample_ids,
                        stringsAsFactors = FALSE)

  if (n_workers > 1L && requireNamespace("parallel", quietly = TRUE)) {
    cl <- parallel::makeCluster(n_workers)
    on.exit(parallel::stopCluster(cl), add = TRUE)
    parallel::clusterExport(
      cl,
      varlist = c("df", "group_col", "hit_markers", "method",
                  "fdr", "neg_frac", "min_pos_rate", "max_pos_rate_cap",
                  "max_pos_rate", "n_grid", "eps",
                  "f_quant_thr", "f_fdr_thr", "f_bfdr_thr",
                  ".get_marker_val", ".clamp_thr", ".enforce_mpr",
                  "compute_group_thr"),
      envir = environment()
    )
    parallel::clusterEvalQ(cl, library(dplyr))
    thr_long <- do.call(
      bind_rows,
      parallel::clusterMap(
        cl,
        compute_group_thr,
        m = groups$marker,
        s = groups$sample
      )
    )
  } else {
    thr_long <- map2_dfr(groups$marker, groups$sample, compute_group_thr)
  }

  # ── Diagnostics: den_at_thr and fallback flag ────────────────────────────────
  # Compute positive call rate at the selected threshold per group,
  # and flag groups where the threshold equals the min_pos_rate quantile floor.
  thr_long <- thr_long %>%
    rowwise() %>%
    mutate(
      den_at_thr = {
        v <- df[[marker]][df[[group_col]] == .data[[group_col]]]
        v <- v[is.finite(v)]
        if (is.finite(thr) && length(v)) mean(v >= thr) else NA_real_
      },
      q_1_minus_mpr = {
        v <- df[[marker]][df[[group_col]] == .data[[group_col]]]
        v <- v[is.finite(v)]
        if (length(v))
          as.numeric(quantile(v, probs = 1 - mpr,
                              na.rm = TRUE, names = FALSE, type = 7))
        else NA_real_
      }
    ) %>%
    ungroup() %>%
    mutate(
      thr_equals_q_1_minus_mpr = if_else(
        is.finite(thr) & is.finite(q_1_minus_mpr),
        abs(thr - q_1_minus_mpr) < 1e-10,
        NA
      )
    )

  # ── Apply thresholds ─────────────────────────────────────────────────────────
  # Vectorised per-marker application avoids dplyr pick() overhead.
  thr_tbl <- thr_long %>%
    select(all_of(group_col), marker, thr) %>%
    pivot_wider(names_from = marker, values_from = thr, names_prefix = "thr__")

  base <- df %>%
    select(any_of(c("ImageID", "SampleID", "SubjectID", "X", "Y")),
           all_of(group_col), all_of(hit_markers))

  out <- base
  thr_idx <- match(base[[group_col]], thr_tbl[[group_col]])

  for (m in hit_markers) {
    thr_vec <- thr_tbl[[paste0("thr__", m)]][thr_idx]
    thr_vec[!is.finite(thr_vec)] <- Inf
    out[[m]] <- as.integer(base[[m]] >= thr_vec)
  }

  out <- select(out,
    any_of(union(c("ImageID", "SampleID", "SubjectID", "X", "Y"), group_col)),
    all_of(hit_markers)
  )

  if (!return_diagnostics) return(out)

  # ── Diagnostics object ───────────────────────────────────────────────────────
  if (!is.null(diagnostics_markers))
    thr_long <- filter(thr_long, marker %in% diagnostics_markers)
  if (!is.null(diagnostics_groups))
    thr_long <- filter(thr_long, .data[[group_col]] %in% diagnostics_groups)

  list(
    out         = out,
    diagnostics = list(
      method      = method,
      group_col   = group_col,
      hit_markers = hit_markers,
      thr_tbl     = thr_tbl,
      thr_long    = thr_long
    )
  )
}

```

## 1d. Run binarization

```{r binarize}
binary_result <- make_marker_binary(
  df               = df_base,
  markers          = names(marker_fdr),
  method           = "fdr_bayesian",
  group_col        = "SampleID",
  fdr              = marker_fdr,
  min_pos_rate     = marker_min_pos,
  max_pos_rate_cap = marker_max_cap,
  return_diagnostics = TRUE
)

spatial_df_univariate <- binary_result$out

cat("Binary calls generated for",
    format(nrow(spatial_df_univariate), big.mark = ","),
    "cells across",
    n_distinct(spatial_df_univariate$SampleID), "samples.\n")
```

## 1e. Threshold summary

The table below shows the per-marker per-sample thresholds that were applied
and the resulting positive call rate. `Fallback = ⚠ Yes` means the Bayesian
FDR constraint was unsatisfiable for that marker × sample and the algorithm
defaulted to the `min_pos_rate` floor — revisit the corresponding `fdr` or
`max_pos_rate_cap` setting if this occurs frequently for a given marker.

```{r threshold-summary, echo=FALSE}
binary_result$diagnostics$thr_long %>%
  mutate(
    thr        = round(thr, 4),
    den_at_thr = round(den_at_thr, 4),
    fallback   = if_else(thr_equals_q_1_minus_mpr == TRUE, "⚠ Yes", "—")
  ) %>%
  select(
    Sample      = SampleID,
    Marker      = marker,
    Threshold   = thr,
    `Pos rate`  = den_at_thr,
    `Fallback?` = fallback
  ) %>%
  kbl(caption = paste0(
    "Applied thresholds and positive call rates per marker × sample. ",
    "⚠ = FDR constraint unsatisfiable; fell back to min_pos_rate floor."
  )) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  scroll_box(width = "100%", height = "500px")
```

---

# 2. Phenotype Assignment

## 2a. Lineage and phenotype rules

**Specify the biological gating hierarchy here.** Two things to set:

**`phenotype_levels`** — a named list defining which phenotypes belong to
each lineage and their display order in plots. Lineage names must match
exactly what the `Lineage` assignment produces below.

**Phenotype `case_when` block** — ordered from most to least specific
within each lineage. The first matching rule wins. Add new phenotypes
above the lineage-level catch-all for that lineage (e.g. above
`Lineage == "T" ~ "T cell (CD3e+)"`). Any marker referenced here must
have been included in the binarization step above.

```{r phenotype-levels}
# ── USER-SPECIFIED ────────────────────────────────────────────────────────────
# Define phenotype membership and display order per lineage.
# Names must match Lineage values produced in the case_when block below.

phenotype_levels <- list(
  "T" = c(
    "T cell (CD3e+)",
    "Helper T (CD3e+CD4+CD8-)",
    "Cytotoxic T (CD3e+CD8+)",
    "Regulatory T (CD3e+CD4+FOXP3+CD39+)"
  ),
  "NK" = c(
    "NK cell (CD3e-CD56+)"
  ),
  "B" = c(
    "B cell (CD20+)"
  ),
  "Myeloid" = c(
    "Dendritic cell (CD11c+HLA-DR+CD3e-CD20-CD68low)",
    "Neutrophil (CD66+MPO+)",
    "M-MDSC (CD11b+CD14+HLA-DRlow)",
    "PMN-MDSC (CD11b+CD14-CD66+MPO+HLA-DRlow)",
    "M1 macrophage (CD68+iNOS+CD163low)",
    "M2 macrophage (CD68+CD163+CD206+)",
    "Pan-macrophage (CD68+)"
  )
)

```

```{r phenotype-assignment}
spatial_df_final <- spatial_df_univariate %>%
  mutate(across(all_of(names(marker_fdr)), function(x) as.integer(coalesce(as.integer(x), 0L)))) %>%
  rename(HLA_DR = `HLA-DR`) %>%
  mutate(

    # ── Step 1: Lineage ───────────────────────────────────────────────────────
    is_T       = CD3e == 1,
    is_B       = CD20 == 1,
    is_NK      = CD3e == 0 & CD56 == 1,
    is_Myeloid = CD11b == 1 | CD68 == 1 | CD66 == 1,
    n_lin      = as.integer(is_T) + as.integer(is_B) +
                 as.integer(is_NK) + as.integer(is_Myeloid),

    Lineage = case_when(
      n_lin >= 2 ~ "Ambiguous/Double+",
      is_T       ~ "T",
      is_B       ~ "B",
      is_NK      ~ "NK",
      is_Myeloid ~ "Myeloid",
      TRUE       ~ "Other"
    ),

    # ── Step 2: Phenotype ─────────────────────────────────────────────────────
    # Rules evaluated top-to-bottom; first match wins.
    # More specific rules must precede less specific ones within each lineage.
    Phenotype = case_when(

      Lineage == "Ambiguous/Double+" ~ "Ambiguous/Double+",

      # T cells
      Lineage == "T" & CD4 == 1 & FOXP3 == 1 & CD39 == 1 ~
        "Regulatory T (CD3e+CD4+FOXP3+CD39+)",
      Lineage == "T" & CD8 == 1 ~
        "Cytotoxic T (CD3e+CD8+)",
      Lineage == "T" & CD4 == 1 & CD8 == 0 ~
        "Helper T (CD3e+CD4+CD8-)",
      Lineage == "T" ~
        "T cell (CD3e+)",

      # NK cells
      Lineage == "NK" ~
        "NK cell (CD3e-CD56+)",

      # B cells
      Lineage == "B" ~
        "B cell (CD20+)",

      # Dendritic cells
      Lineage == "Myeloid" & CD11c == 1 & HLA_DR == 1 & CD3e == 0 & CD20 == 0 & CD68 == 0 ~
        "Dendritic cell (CD11c+HLA-DR+CD3e-CD20-CD68low)",

      # Neutrophils
      Lineage == "Myeloid" & CD66 == 1 & MPO == 1 ~
        "Neutrophil (CD66+MPO+)",

      # MDSCs
      Lineage == "Myeloid" & CD11b == 1 & CD14 == 1 & HLA_DR == 0 ~
        "M-MDSC (CD11b+CD14+HLA-DRlow)",
      Lineage == "Myeloid" & CD11b == 1 & CD14 == 0 & CD66 == 1 & MPO == 1 & HLA_DR == 0 ~
        "PMN-MDSC (CD11b+CD14-CD66+MPO+HLA-DRlow)",

      # Macrophages
      Lineage == "Myeloid" & CD68 == 1 & iNOS == 1 & CD163 == 0 ~
        "M1 macrophage (CD68+iNOS+CD163low)",
      Lineage == "Myeloid" & CD68 == 1 & CD163 == 1 & CD206 == 1 ~
        "M2 macrophage (CD68+CD163+CD206+)",
      Lineage == "Myeloid" & CD68 == 1 ~
        "Pan-macrophage (CD68+)",

      TRUE ~ "Other/Unassigned"
    )
  ) %>%
  select(-is_T, -is_B, -is_NK, -is_Myeloid, -n_lin)

```

## 2b. Assignment summary

```{r assignment-summary, echo=FALSE}

spatial_df_final %>%
  count(Lineage) %>%
  mutate(`%` = round(n / sum(n) * 100, 2)) %>%
  arrange(desc(n)) %>%
  kbl(caption = "Lineage assignment summary across all cells and samples") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE)

```

```{r phenotype-summary-table, echo=FALSE}

spatial_df_final %>%
  filter(!Phenotype %in% c("Other/Unassigned", "Ambiguous/Double+")) %>%
  count(Lineage, Phenotype) %>%
  group_by(Lineage) %>%
  mutate(`% of lineage` = round(n / sum(n) * 100, 2)) %>%
  ungroup() %>%
  arrange(Lineage, desc(n)) %>%
  kbl(caption = "Phenotype composition within each lineage (excluding Other/Unassigned and Ambiguous)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  scroll_box(width = "100%")

```

---

# 3. Gating and Phenotyping QC

```{r phenotyping-qc-fn, include=FALSE}

phenotyping_qc_report <- function(
  binary_result,
  spatial_df_final,
  markers,
  phenotype_levels  = NULL,
  sample_col        = "SampleID",
  lineage_col       = "Lineage",
  phenotype_col     = "Phenotype",
  pos_rate_flag_lo  = 0.001,
  pos_rate_flag_hi  = 0.50,
  fallback_flag_pct = 0.25
) {

  if (!is.list(binary_result) || !"diagnostics" %in% names(binary_result))
    stop("binary_result must be output of make_marker_binary(..., return_diagnostics = TRUE).")

  missing_cols <- setdiff(c(sample_col, lineage_col, phenotype_col),
                          names(spatial_df_final))
  if (length(missing_cols) > 0)
    stop("spatial_df_final missing column(s): ",
         paste(missing_cols, collapse = ", "))

  diag     <- binary_result$diagnostics
  thr_long <- diag$thr_long
  df       <- spatial_df_final
  n_cells  <- nrow(df)

  # ── Section 1: Gating diagnostics ───────────────────────────────────────────
  gating_diag <- thr_long %>%
    group_by(marker) %>%
    summarise(
      n_samples     = n(),
      n_fallback    = sum(thr_equals_q_1_minus_mpr == TRUE, na.rm = TRUE),
      pct_fallback  = round(n_fallback / n_samples * 100, 1),
      mean_thr      = round(mean(thr, na.rm = TRUE), 4),
      sd_thr        = round(sd(thr,  na.rm = TRUE), 4),
      min_thr       = round(min(thr, na.rm = TRUE), 4),
      max_thr       = round(max(thr, na.rm = TRUE), 4),
      mean_pos_rate = round(mean(den_at_thr, na.rm = TRUE), 4),
      sd_pos_rate   = round(sd(den_at_thr,   na.rm = TRUE), 4),
      .groups       = "drop"
    ) %>%
    mutate(
      flag_high_fallback = pct_fallback > fallback_flag_pct * 100,
      flag_unstable_thr  = sd_thr > 0.15,
      interpretation = case_when(
        flag_high_fallback & flag_unstable_thr ~
          "High fallback + unstable threshold: FDR likely too strict or cap too low",
        flag_high_fallback ~
          "High fallback: FDR unsatisfiable in many samples; relax fdr or raise cap",
        flag_unstable_thr ~
          "High threshold variance: heterogeneous staining; review per-sample violins",
        TRUE ~ "OK"
      )
    ) %>%
    arrange(desc(pct_fallback))

  gating_diag_per_sample <- thr_long %>%
    filter(thr_equals_q_1_minus_mpr == TRUE) %>%
    select(sample = all_of(diag$group_col), marker,
           thr, den_at_thr, q_1_minus_mpr) %>%
    arrange(marker, sample)

  # ── Section 2: Positive rate audit ──────────────────────────────────────────
  hit_markers <- intersect(markers, names(df))

  pos_rate_audit <- df %>%
    select(all_of(c(sample_col, hit_markers))) %>%
    pivot_longer(cols = all_of(hit_markers),
                 names_to = "marker", values_to = "call") %>%
    group_by(.data[[sample_col]], marker) %>%
    summarise(
      n_cells    = n(),
      n_positive = sum(call == 1L, na.rm = TRUE),
      pos_rate   = round(n_positive / n_cells, 5),
      .groups    = "drop"
    ) %>%
    mutate(
      flag_lo = pos_rate < pos_rate_flag_lo,
      flag_hi = pos_rate > pos_rate_flag_hi,
      flag    = flag_lo | flag_hi,
      note    = case_when(
        flag_lo ~ paste0("Positivity < ", pos_rate_flag_lo * 100,
                         "% — possible staining failure or over-strict threshold"),
        flag_hi ~ paste0("Positivity > ", pos_rate_flag_hi * 100,
                         "% — possible non-specific staining or threshold too permissive"),
        TRUE    ~ "OK"
      )
    ) %>%
    arrange(desc(flag), marker, .data[[sample_col]])

  pos_rate_summary <- pos_rate_audit %>%
    group_by(marker) %>%
    summarise(
      mean_pos_rate = round(mean(pos_rate), 5),
      sd_pos_rate   = round(sd(pos_rate),   5),
      min_pos_rate  = round(min(pos_rate),   5),
      max_pos_rate  = round(max(pos_rate),   5),
      n_flag_lo     = sum(flag_lo),
      n_flag_hi     = sum(flag_hi),
      n_flagged     = sum(flag),
      .groups       = "drop"
    ) %>%
    arrange(desc(n_flagged))

  # ── Section 3: Phenotype rule audit (pre-hierarchy) ─────────────────────────
  # Rules are derived from phenotype_levels: each named phenotype string is
  # matched against the assigned Phenotype column independently, ignoring
  # case_when priority order.
  rule_audit <- NULL

  if (!is.null(phenotype_levels)) {
    all_phenotypes <- unlist(phenotype_levels, use.names = FALSE)
    rule_audit <- map_dfr(all_phenotypes, function(ph) {
      n_satisfy <- sum(df[[phenotype_col]] == ph, na.rm = TRUE)
      tibble(
        rule        = ph,
        n_satisfy   = n_satisfy,
        pct_satisfy = round(n_satisfy / n_cells * 100, 4)
      )
    }) %>%
      arrange(desc(n_satisfy))
  }

  # ── Section 4: Rule collision audit ─────────────────────────────────────────
  # Among cells assigned to each phenotype, what fraction were also the
  # only-possible assignment (i.e. did the hierarchy resolve ambiguity)?
  # This uses the lineage breakdown to surface where hierarchy does heavy lifting.
  collision_audit <- NULL

  if (!is.null(phenotype_levels)) {
    all_phenotypes <- unlist(phenotype_levels, use.names = FALSE)

    collision_audit <- map_dfr(all_phenotypes, function(ph) {
      n_assigned <- sum(df[[phenotype_col]] == ph, na.rm = TRUE)
      if (n_assigned == 0)
        return(tibble(rule = ph, collides_with = "none",
                      n_collision = 0L, pct_of_rule = 0))
      others <- setdiff(all_phenotypes, ph)
      map_dfr(others, function(other) {
        n_col <- sum(df[[phenotype_col]] == ph &
                     df[[lineage_col]] == df[[lineage_col]], na.rm = TRUE)
        tibble(
          rule          = ph,
          collides_with = other,
          n_collision   = sum(
            df[[phenotype_col]] == ph &
              df[[lineage_col]] %in%
              names(which(sapply(phenotype_levels,
                                 function(x) other %in% x))),
            na.rm = TRUE
          ),
          pct_of_rule = round(n_collision / n_assigned * 100, 2)
        )
      })
    }) %>%
      filter(n_collision > 0) %>%
      arrange(desc(pct_of_rule))
  }

  # ── Section 5: Lineage breakdown ─────────────────────────────────────────────
  lineage_breakdown <- df %>%
    group_by(.data[[sample_col]], .data[[lineage_col]]) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(.data[[sample_col]]) %>%
    mutate(total_cells = sum(n), pct = round(n / total_cells * 100, 2)) %>%
    ungroup() %>%
    arrange(.data[[sample_col]], desc(n))

  lineage_summary <- lineage_breakdown %>%
    group_by(.data[[lineage_col]]) %>%
    summarise(
      mean_pct = round(mean(pct), 2), sd_pct = round(sd(pct), 2),
      min_pct  = round(min(pct),  2), max_pct = round(max(pct), 2),
      .groups  = "drop"
    ) %>%
    arrange(desc(mean_pct))

  unassigned_flag <- lineage_breakdown %>%
    filter(.data[[lineage_col]] %in% c("Other", "Other/Unassigned")) %>%
    filter(pct > 50) %>%
    select(all_of(sample_col), pct_unassigned = pct)

  # ── Section 6: Phenotype breakdown ───────────────────────────────────────────
  phenotype_breakdown <- df %>%
    filter(!.data[[phenotype_col]] %in%
             c("Other/Unassigned", "Ambiguous/Double+")) %>%
    group_by(.data[[sample_col]], .data[[lineage_col]],
             .data[[phenotype_col]]) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(.data[[sample_col]], .data[[lineage_col]]) %>%
    mutate(lineage_total  = sum(n),
           pct_of_lineage = round(n / lineage_total * 100, 2)) %>%
    ungroup() %>%
    arrange(.data[[sample_col]], .data[[lineage_col]], desc(n))

  phenotype_summary <- phenotype_breakdown %>%
    group_by(.data[[lineage_col]], .data[[phenotype_col]]) %>%
    summarise(
      mean_pct = round(mean(pct_of_lineage), 2),
      sd_pct   = round(sd(pct_of_lineage),   2),
      min_pct  = round(min(pct_of_lineage),   2),
      max_pct  = round(max(pct_of_lineage),   2),
      .groups  = "drop"
    ) %>%
    arrange(.data[[lineage_col]], desc(mean_pct))

  # ── Section 7: Ambiguous / fallthrough ───────────────────────────────────────
  fallthrough_summary <- df %>%
    group_by(.data[[sample_col]]) %>%
    summarise(
      n_total        = n(),
      n_ambiguous    = sum(.data[[phenotype_col]] == "Ambiguous/Double+",
                           na.rm = TRUE),
      n_unassigned   = sum(.data[[phenotype_col]] == "Other/Unassigned",
                           na.rm = TRUE),
      pct_ambiguous  = round(n_ambiguous  / n_total * 100, 2),
      pct_unassigned = round(n_unassigned / n_total * 100, 2),
      .groups        = "drop"
    ) %>%
    mutate(
      flag_ambiguous  = pct_ambiguous  > 5,
      flag_unassigned = pct_unassigned > 30,
      note = case_when(
        flag_ambiguous & flag_unassigned ~
          "High ambiguous + high unassigned: review lineage caps and phenotype rule coverage",
        flag_ambiguous  ~
          "High ambiguous: lineage caps may be too permissive",
        flag_unassigned ~
          "High unassigned: phenotype rules do not cover many cells in this sample",
        TRUE ~ "OK"
      )
    ) %>%
    arrange(desc(pct_unassigned))

  invisible(list(
    gating_diag            = gating_diag,
    gating_diag_per_sample = gating_diag_per_sample,
    pos_rate_audit         = pos_rate_audit,
    pos_rate_summary       = pos_rate_summary,
    rule_audit             = rule_audit,
    collision_audit        = collision_audit,
    lineage_breakdown      = lineage_breakdown,
    lineage_summary        = lineage_summary,
    phenotype_breakdown    = phenotype_breakdown,
    phenotype_summary      = phenotype_summary,
    fallthrough_summary    = fallthrough_summary
  ))
}

```

```{r run-qc-report}

qc_report <- phenotyping_qc_report(
  binary_result    = binary_result,
  spatial_df_final = spatial_df_final,
  markers          = names(marker_fdr),
  phenotype_levels = phenotype_levels,
  sample_col       = sample_col,
  lineage_col      = lineage_col,
  phenotype_col    = phenotype_col
)

```

## 3a. Gating diagnostics

A **fallback** occurs when the Bayesian FDR constraint is unsatisfiable
for a given marker × sample combination and the threshold defaults to the
`min_pos_rate` quantile floor. High fallback rates suggest `fdr` is too
strict or `max_pos_rate_cap` is too low for that marker. **Threshold
instability** (high SD across samples) suggests heterogeneous staining or
genuine biological variability across the cohort.

```{r gating-diag-table, echo=FALSE}

qc_report$gating_diag %>%
  mutate(
    pct_fallback       = paste0(pct_fallback, "%"),
    flag_high_fallback = if_else(flag_high_fallback, "\u26a0 Yes", "\u2014"),
    flag_unstable_thr  = if_else(flag_unstable_thr,  "\u26a0 Yes", "\u2014")
  ) %>%
  select(
    Marker           = marker,
    `Samples (n)`    = n_samples,
    `Fallback (%)`   = pct_fallback,
    `High fallback?` = flag_high_fallback,
    `Mean thr`       = mean_thr,
    `SD thr`         = sd_thr,
    `Mean pos rate`  = mean_pos_rate,
    Interpretation   = interpretation
  ) %>%
  kbl(caption = "Per-marker gating diagnostics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(3, bold = TRUE,
              color = if_else(qc_report$gating_diag$flag_high_fallback,
                              FLAG_RED, "black")) %>%
  column_spec(5, color = if_else(qc_report$gating_diag$flag_unstable_thr,
                                 FLAG_ORANGE, "black")) %>%
  scroll_box(width = "100%")

```

```{r gating-diag-plot, echo=FALSE, fig.height=5}

binary_result$diagnostics$thr_long %>%
  ggplot(aes(x = reorder(marker, thr, median), y = thr)) +
  geom_boxplot(outlier.shape = NA, fill = NEUTRAL, linewidth = 0.5) +
  geom_jitter(aes(color = thr_equals_q_1_minus_mpr),
              width = 0.15, size = 1.5, alpha = 0.7) +
  scale_color_manual(
    values   = c("TRUE" = FLAG_RED, "FALSE" = "grey40"),
    na.value = "grey80",
    labels   = c("TRUE" = "Fallback", "FALSE" = "FDR-solved"),
    name     = "Threshold type"
  ) +
  labs(
    title    = "Threshold distribution by marker across samples",
    subtitle = "Red = fallback to min_pos_rate floor",
    x = "Marker", y = "Threshold value"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

```

```{r gating-fallback-detail, echo=FALSE}

if (nrow(qc_report$gating_diag_per_sample) > 0) {
  qc_report$gating_diag_per_sample %>%
    rename(Sample = sample, Marker = marker, Threshold = thr,
           `Pos rate` = den_at_thr, `MPR quantile` = q_1_minus_mpr) %>%
    kbl(caption = "Per-sample fallback details (fallback cases only)") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE) %>%
    scroll_box(width = "100%", height = "400px")
} else {
  cat("\u2713 No fallback cases detected across any marker \u00d7 sample combination.\n")
}

```

## 3b. Positive rate audit

Positivity rates are the fraction of cells called positive per marker per
sample. Unusually low rates suggest staining failure or an over-strict
threshold. Unusually high rates suggest non-specific staining or an overly
permissive threshold.

```{r pos-rate-summary, echo=FALSE}

qc_report$pos_rate_summary %>%
  rename(Marker = marker, `Mean pos rate` = mean_pos_rate,
         SD = sd_pos_rate, Min = min_pos_rate, Max = max_pos_rate,
         `N flag low` = n_flag_lo, `N flag high` = n_flag_hi,
         `N flagged` = n_flagged) %>%
  kbl(caption = "Per-marker positivity rate summary across samples") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(8, bold = TRUE,
              color = if_else(qc_report$pos_rate_summary$n_flagged > 0,
                              FLAG_RED, "black")) %>%
  scroll_box(width = "100%")

```

```{r pos-rate-heatmap, echo=FALSE, fig.height=6}

qc_report$pos_rate_audit %>%
  mutate(pos_rate_pct = pos_rate * 100) %>%
  ggplot(aes(x = .data[[sample_col]], y = marker, fill = pos_rate_pct)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = if_else(flag, "\u26a0", "")),
            size = 3, color = "black") +
  scale_fill_viridis_c(option = "magma", direction = -1,
                       name = "Positivity (%)", limits = c(0, NA)) +
  labs(title    = "Marker positivity rates by sample",
       subtitle = "\u26a0 = flagged (unexpectedly low or high)",
       x = "Sample", y = "Marker") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, size = 8),
        panel.grid  = element_blank())

```

```{r pos-rate-flagged, echo=FALSE}

flagged_pos <- filter(qc_report$pos_rate_audit, flag)
if (nrow(flagged_pos) > 0) {
  flagged_pos %>%
    select(Sample = all_of(sample_col), Marker = marker,
           `N cells` = n_cells, `N positive` = n_positive,
           `Pos rate` = pos_rate, Note = note) %>%
    kbl(caption = "Flagged marker \u00d7 sample combinations") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE) %>%
    scroll_box(width = "100%", height = "400px")
} else {
  cat("\u2713 No marker \u00d7 sample combinations flagged.\n")
}

```

## 3c. Phenotype rule audit

Each phenotype label is evaluated against the assigned `Phenotype` column
independently, showing how many cells received that label. This makes visible
where the `case_when` hierarchy is doing heavy lifting — a phenotype with very
few assigned cells but a biologically expected prevalence may indicate a rule
conflict or ordering issue.

```{r rule-audit, echo=FALSE}

if (!is.null(qc_report$rule_audit)) {
  qc_report$rule_audit %>%
    rename(Phenotype = rule, `N assigned` = n_satisfy,
           `% of all cells` = pct_satisfy) %>%
    kbl(caption = "Phenotype assignment counts") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE) %>%
    column_spec(2, bold = TRUE) %>%
    scroll_box(width = "100%")
} else {
  cat("Skipped \u2014 pass `phenotype_levels` to enable.\n")
}

```

```{r rule-audit-plot, echo=FALSE, fig.height=5}

if (!is.null(qc_report$rule_audit)) {
  qc_report$rule_audit %>%
    mutate(rule = fct_reorder(rule, pct_satisfy)) %>%
    ggplot(aes(x = rule, y = pct_satisfy)) +
    geom_col(fill = NEUTRAL, color = "grey40", linewidth = 0.3) +
    geom_text(aes(label = paste0(round(pct_satisfy, 3), "%")),
              hjust = -0.1, size = 3.2) +
    coord_flip(clip = "off") +
    expand_limits(y = max(qc_report$rule_audit$pct_satisfy) * 1.25) +
    labs(title = "Phenotype assignment rates",
         x = NULL, y = "% of all cells") +
    theme_bw(base_size = 12)
}

```

## 3d. Lineage breakdown

```{r lineage-summary, echo=FALSE}

qc_report$lineage_summary %>%
  rename(Lineage = all_of(lineage_col), `Mean %` = mean_pct,
         `SD %` = sd_pct, `Min %` = min_pct, `Max %` = max_pct) %>%
  kbl(caption = "Lineage composition summary across samples") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

```

```{r lineage-stacked-bar, echo=FALSE, fig.height=6}

qc_report$lineage_breakdown %>%
  mutate(!!sample_col := fct_reorder(
    .data[[sample_col]],
    if_else(.data[[lineage_col]] == "Other", pct, 0),
    .fun = sum
  )) %>%
  ggplot(aes(x = .data[[sample_col]], y = pct,
             fill = .data[[lineage_col]])) +
  geom_col(position = "stack", linewidth = 0.2, color = "white") +
  scale_fill_brewer(palette = "Set2", name = "Lineage") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title    = "Lineage composition per sample",
       subtitle = "Ordered by Other fraction",
       x = "Sample", y = "% of total cells") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))

```

```{r lineage-boxplot, echo=FALSE, fig.height=4}

qc_report$lineage_breakdown %>%
  ggplot(aes(x = reorder(.data[[lineage_col]], pct, median), y = pct)) +
  geom_boxplot(fill = NEUTRAL, outlier.shape = NA, linewidth = 0.5) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6, color = "grey30") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Lineage abundance distribution across samples",
       x = NULL, y = "% of total cells") +
  theme_bw(base_size = 12) +
  coord_flip()

```

## 3e. Phenotype breakdown

```{r phenotype-breakdown-table, echo=FALSE}

qc_report$phenotype_summary %>%
  rename(Lineage = all_of(lineage_col), Phenotype = all_of(phenotype_col),
         `Mean % of lineage` = mean_pct, `SD %` = sd_pct,
         `Min %` = min_pct, `Max %` = max_pct) %>%
  kbl(caption = "Phenotype composition within lineage, averaged across samples") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  scroll_box(width = "100%")

```

```{r phenotype-breakdown-plot, echo=FALSE, fig.height=8}

qc_report$phenotype_breakdown %>%
  mutate(Phenotype = str_wrap(.data[[phenotype_col]], width = 35)) %>%
  ggplot(aes(x = reorder(Phenotype, pct_of_lineage, median),
             y = pct_of_lineage)) +
  geom_boxplot(fill = NEUTRAL, outlier.shape = NA, linewidth = 0.4) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.55, color = "grey30") +
  facet_wrap(vars(.data[[lineage_col]]), scales = "free", ncol = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Phenotype abundance within lineage across samples",
       x = NULL, y = "% of lineage cells") +
  theme_bw(base_size = 11) +
  theme(axis.text.y = element_text(size = 8)) +
  coord_flip()

```

## 3f. Ambiguous and unassigned cells

Cells labeled `Ambiguous/Double+` satisfied multiple lineage rules
simultaneously. Cells labeled `Other/Unassigned` passed lineage assignment
but matched no phenotype rule. High rates of either suggest lineage cap
settings or phenotype rule gaps that need attention.

```{r fallthrough-table, echo=FALSE}

qc_report$fallthrough_summary %>%
  select(Sample = all_of(sample_col), `N total` = n_total,
         `N ambiguous` = n_ambiguous, `% ambiguous` = pct_ambiguous,
         `N unassigned` = n_unassigned, `% unassigned` = pct_unassigned,
         `Flag ambiguous` = flag_ambiguous,
         `Flag unassigned` = flag_unassigned, Note = note) %>%
  mutate(`Flag ambiguous`  = if_else(`Flag ambiguous`,  "\u26a0 Yes", "\u2014"),
         `Flag unassigned` = if_else(`Flag unassigned`, "\u26a0 Yes", "\u2014")) %>%
  kbl(caption = "Ambiguous and unassigned cell rates per sample") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(4, color = if_else(qc_report$fallthrough_summary$flag_ambiguous,
                                 FLAG_RED, "black")) %>%
  column_spec(6, color = if_else(qc_report$fallthrough_summary$flag_unassigned,
                                 FLAG_RED, "black")) %>%
  scroll_box(width = "100%")

```

```{r fallthrough-plot, echo=FALSE, fig.height=5}

qc_report$fallthrough_summary %>%
  pivot_longer(cols = c(pct_ambiguous, pct_unassigned),
               names_to = "type", values_to = "pct") %>%
  mutate(
    type   = recode(type, pct_ambiguous  = "Ambiguous/Double+",
                          pct_unassigned = "Other/Unassigned"),
    sample = fct_reorder(.data[[sample_col]], pct, max)
  ) %>%
  ggplot(aes(x = sample, y = pct, fill = type)) +
  geom_col(position = "dodge", linewidth = 0.3, color = "white") +
  scale_fill_manual(values = c("Ambiguous/Double+" = FLAG_ORANGE,
                               "Other/Unassigned"  = FLAG_RED),
                    name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title    = "Ambiguous and unassigned rates per sample",
       subtitle = "Ordered by maximum of the two rates",
       x = "Sample", y = "% of total cells") +
  theme_bw(base_size = 12) +
  theme(axis.text.x   = element_text(angle = 50, hjust = 1),
        legend.position = "top")

```

---

# 4. Abundance Summary

Per-sample phenotype proportions out of total cells (including
`Other/Unassigned` and `Ambiguous/Double+` in the denominator). These are
the primary inputs to downstream statistical analysis.

```{r prepare-phenotype-summary-fn, include=FALSE}

prepare_phenotype_summary_from_image <- function(
  df,
  cohort_patterns    = NULL,
  image_col          = "ImageID",
  sample_col         = "SampleID",
  lineage_col        = "Lineage",
  exclude_phenotypes = c("Other/Unassigned", "Ambiguous/Double+"),
  normalize_within   = NULL,
  phenotype_levels   = NULL,
  lineage_levels     = NULL,
  cohort_levels      = NULL,
  ignore_case        = TRUE,
  warn_parse         = TRUE,
  cohort_id_col      = "ImageID"
) {
  missing_cols <- setdiff(c(image_col, sample_col, lineage_col, "Phenotype"),
                          names(df))
  if (length(missing_cols) > 0)
    stop("`df` is missing required column(s): ",
         paste(missing_cols, collapse = ", "))
  if (!cohort_id_col %in% names(df))
    stop(sprintf("cohort_id_col = '%s' not found in `df`.", cohort_id_col))

  df <- mutate(df, across(
    all_of(c(image_col, sample_col, lineage_col, "Phenotype", cohort_id_col)),
    as.character
  ))

  # Denominators computed before any filtering
  totals_df <- df %>%
    group_by(.data[[sample_col]]) %>%
    summarise(total_cells = n(), .groups = "drop")

  df <- if (!is.null(cohort_patterns) && length(cohort_patterns) > 0) {
    pats         <- as.character(cohort_patterns)
    cohort_names <- if (is.null(names(cohort_patterns)) ||
                        any(names(cohort_patterns) == ""))
      pats else names(cohort_patterns)
    parse_one <- function(z) {
      if (is.na(z) || !nzchar(z)) return(NA_character_)
      hit <- vapply(pats, function(p)
        grepl(p, z, ignore.case = ignore_case, perl = TRUE), logical(1))
      if (any(hit)) cohort_names[which(hit)[1]] else NA_character_
    }
    out <- mutate(df, Cohort = vapply(.data[[cohort_id_col]],
                                      parse_one, character(1)))
    if (warn_parse) {
      unmatched <- out %>%
        filter(is.na(Cohort)) %>%
        distinct(.data[[cohort_id_col]]) %>%
        pull(.data[[cohort_id_col]])
      if (length(unmatched) > 0)
        warning(sprintf(
          "Cohort parsing: %d unique `%s` value(s) had no match: %s%s",
          length(unmatched), cohort_id_col,
          paste(head(unmatched, 10), collapse = ", "),
          if (length(unmatched) > 10) " ..." else ""
        ), call. = FALSE)
    }
    out
  } else if ("Cohort" %in% names(df)) {
    mutate(df, Cohort = as.character(Cohort))
  } else {
    mutate(df, Cohort = "All samples")
  }

  df <- filter(df, !is.na(Cohort), nzchar(Cohort))
  if (nrow(df) == 0) stop("No rows remain after cohort parsing/filtering.")

  observed_cohorts <- unique(df$Cohort)

  if (!is.null(cohort_levels) && length(cohort_levels) > 0) {
    valid_levels <- intersect(cohort_levels, observed_cohorts)
    if (length(valid_levels) == 0)
      stop("`cohort_levels` has no overlap with parsed cohorts.\n",
           "Parsed: ",   paste(observed_cohorts, collapse = ", "), "\n",
           "Supplied: ", paste(cohort_levels,    collapse = ", "))
    dropped <- setdiff(observed_cohorts, valid_levels)
    if (warn_parse && length(dropped) > 0)
      warning("Dropping cohort(s) not in `cohort_levels`: ",
              paste(dropped, collapse = ", "), call. = FALSE)
    df            <- filter(df, Cohort %in% valid_levels)
    cohort_levels <- valid_levels
  } else {
    cohort_levels <- observed_cohorts
  }

  df <- mutate(df, Cohort = factor(Cohort, levels = cohort_levels))

  df_f <- df %>%
    filter(!is.na(.data[[lineage_col]]), nzchar(.data[[lineage_col]]),
           !is.na(Phenotype), nzchar(Phenotype),
           !Phenotype %in% exclude_phenotypes)

  if (nrow(df_f) == 0) stop("No rows remain after lineage/phenotype filtering.")

  expected_phenotypes <- if (!is.null(phenotype_levels) &&
                              length(phenotype_levels) > 0) {
    lapply(phenotype_levels,
           function(v) unique(na.omit(as.character(v))))
  } else {
    tmp <- df_f %>%
      distinct(.data[[lineage_col]], Phenotype) %>%
      arrange(.data[[lineage_col]], Phenotype)
    split(as.character(tmp$Phenotype), tmp[[lineage_col]])
  }

  expected_phenotypes <- expected_phenotypes[lengths(expected_phenotypes) > 0]

  lineage_levels <- if (is.null(lineage_levels) || length(lineage_levels) == 0)
    names(expected_phenotypes)
  else
    intersect(lineage_levels, names(expected_phenotypes))

  if (length(lineage_levels) == 0)
    stop("No lineage levels remain after intersecting with observed lineages.")

  expected_phenotypes <- expected_phenotypes[lineage_levels]

  lineage_pheno_tbl <- imap_dfr(
    expected_phenotypes,
    ~ tibble(!!lineage_col := .y, Phenotype = .x, .ord = seq_along(.x))
  ) %>%
    arrange(factor(.data[[lineage_col]], levels = lineage_levels), .ord) %>%
    select(-.ord)

  pheno_levels_global <- unique(lineage_pheno_tbl$Phenotype)

  counts <- df_f %>%
    count(.data[[sample_col]], Cohort, .data[[lineage_col]],
          Phenotype, name = "n")

  sample_cohort <- distinct(df, .data[[sample_col]], Cohort)

  x <- tidyr::crossing(sample_cohort, lineage_pheno_tbl) %>%
    left_join(counts, by = c(sample_col, "Cohort", lineage_col, "Phenotype")) %>%
    mutate(n = tidyr::replace_na(n, 0L)) %>%
    left_join(totals_df, by = sample_col) %>%
    mutate(prop_total = if_else(total_cells > 0, n / total_cells, 0))

  if (!is.null(normalize_within) && length(normalize_within) > 0) {
    normalize_within <- intersect(as.character(normalize_within), lineage_levels)
    x <- x %>%
      group_by(.data[[sample_col]]) %>%
      mutate(
        .within_denom = sum(
          n[as.character(.data[[lineage_col]]) %in% normalize_within],
          na.rm = TRUE
        ),
        prop_within = if_else(
          as.character(.data[[lineage_col]]) %in% normalize_within &
            .within_denom > 0,
          n / .within_denom,
          NA_real_
        )
      ) %>%
      ungroup() %>%
      select(-.within_denom)
  }

  x %>%
    mutate(
      Cohort        = factor(as.character(Cohort), levels = cohort_levels),
      !!lineage_col := factor(as.character(.data[[lineage_col]]),
                              levels = lineage_levels),
      Phenotype     = factor(as.character(Phenotype),
                             levels = pheno_levels_global)
    ) %>%
    arrange(.data[[sample_col]], Cohort,
            .data[[lineage_col]], Phenotype) %>%
    select(all_of(sample_col), Cohort, all_of(lineage_col), Phenotype,
           n, total_cells, starts_with("prop_"))
}

```

```{r abundance}

abundance_plot_df <- spatial_df_final %>%
  prepare_phenotype_summary_from_image(
    cohort_patterns  = NULL,
    cohort_id_col    = "SubjectID",
    image_col        = "ImageID",
    sample_col       = "SampleID",
    lineage_col      = "Lineage",
    phenotype_levels = phenotype_levels,
    lineage_levels   = names(phenotype_levels)
  )

cat("abundance_plot_df:",
    format(nrow(abundance_plot_df), big.mark = ","), "rows\n")
cat("Samples:", n_distinct(abundance_plot_df$SampleID), "\n")
cat("Phenotypes:", n_distinct(abundance_plot_df$Phenotype), "\n")

```

---

# 5. Save Outputs

```{r save}

out_dir <- file.path(dirname(cfg_path), cfg$paths$folders$results)

fwrite(
  spatial_df_final,
  file.path(out_dir, "spatial_phenotype_assignment.csv")
)

fwrite(
  abundance_plot_df,
  file.path(out_dir, "abundance_plot_data.csv")
)

saveRDS(
  phenotype_levels,
  file.path(out_dir, "phenotype_levels.rds")
)

cat("Saved:\n")
cat("  spatial_phenotype_assignment.csv\n")
cat("  abundance_plot_data.csv\n")
cat("  phenotype_levels.rds\n")
cat("Output directory:", out_dir, "\n")

```

---

# Session Information

```{r session}

sessionInfo()

```
