---
title: "PCF Pilot Results: Brooke Fridley"
subtitle: "Fridley Lab"
author: "Kirill Sabitov"
date: today
date-format: long
execute:
  echo: true
  eval: true
  warning: false
  error: false
  message: false
format:
    pdf:
      toc: true
      echo: false
      warning: false
      error: false
      message: false
    html:
      toc: true
      toc-depth: 5
      toc-location: left
      toc-title: "Report Contents"
      number-sections: false
      embed-resources: true
      code-link: true
      code-fold: true
      code-summary: "Show the code"
      page-layout: full
      df-print: paged
output: html_document
---

```{r Packages and Data, echo = F, include = F}

# devtools::install_github("JiangmeiRubyXiong/cfGMM")
# devtools::install_github("JiangmeiRubyXiong/GammaGateR", build_vignettes = FALSE)

.libPaths("/mnt/project/blfridley/PCF_pilots/_common/R/renv/library/linux-rhel-8.10/R-4.4/x86_64-pc-linux-gnu")
options(gt.markdown_engine = "markdown")

suppressPackageStartupMessages({
  library(pacman); library(devtools)
  library(yaml)
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(tidyr)
  library(gtsummary)
  library(ggpubr)
  library(forcats)
  library(rstatix)
  library(kableExtra)
  library(htmltools)
  library(htmlwidgets)
  library(stringi)
  library(stringr)
  library(rlang)
  library(gt)
  library(rmarkdown)
  library(spatialTIME)
  library(purrr)
  library(stringr)
  library(tibble)
  library(scales)
  library(skimr)
  library(purrr)
  library(knitr)
  library(grid)
  library(patchwork)
  library(cowplot)
  library(RColorBrewer)
  library(scattermore)
  library(sass)
  library(cards)
  library(markdown)
  library(commonmark)
  library(litedown)
  library(ComplexUpset)
  library(treemapify)
})

cfg_path <- normalizePath("../config.yml", mustWork = T)
cfg <- read_yaml(cfg_path)

spatial_df_final <- fread(file.path(dirname(cfg_path), cfg$paths$folders$results, "spatial_phenotype_assignment.csv")) %>% 
  as.data.frame()

abundance_plot_df <- fread(file.path(dirname(cfg_path), cfg$paths$folders$results, "abundance_plot_data.csv")) %>% 
  as.data.frame()

phenotype_levels <- readRDS(file.path(dirname(cfg_path), cfg$paths$folders$results, "phenotype_levels.rds"))

metadata <- readxl::read_excel({
  f <- fs::dir_ls(
    file.path(dirname(cfg_path), cfg$paths$folders$pcf_images),
    recurse = FALSE,
    type = "file",
    regexp = "(?i)\\.(xlsx|xls)$"
  )
  f <- f[!stringr::str_detect(basename(f), "^~\\$")]
  f <- f[stringr::str_detect(
    basename(f),
    stringr::regex(cfg$metadata$pattern %||% "PCF_meta_data", ignore_case = TRUE)
  )]
  if (length(f) != 1)
    stop(if (length(f) == 0)
      "No Excel file found matching metadata pattern."
    else
      "Multiple Excel files found matching metadata pattern.")
  f
})

color_vec <- if("Class" %in% names(metadata)){
  setNames(metadata$Class, metadata$SampleID)
} else {
  NULL
}

cohort_levels <- sapply(cfg$cohorts$groups, function(x) x$label)

```

```{r Functions, echo = F, include = F}

## These functions are defined below to make various visualizations throughout the report. 
## They are designed to be generalizable, but you may edit them depending on the pilot's needs. 
## Recommended that you "fold" this chunk (press triangle next to line number) unless you need to edit code. 

plot_lineage <- function(
  x_df,
  lineage_label            = NULL,
  ncol                     = 3,
  add_tests                = TRUE,
  point_alpha              = 0.6,
  point_size               = 1.6,
  box_width                = 0.65,
  pairwise_if_omnibus_sig  = TRUE,
  omnibus_alpha            = 0.05,
  subtitle                 = NULL,
  theme_base               = theme_minimal(base_size = 11),
  phenotype_levels         = NULL,
  cohort_levels            = NULL,
  statisticallycorrect     = TRUE,
  show_ns                  = TRUE,
  color_by                 = NULL,
  color_by_id_col          = "SampleID"
) {

  # ── Validation ───────────────────────────────────────────────────────────────
  missing_cols <- setdiff(c("Lineage", "Phenotype", "prop_total"), names(x_df))
  if (length(missing_cols) > 0)
    stop("Missing required column(s): ", paste(missing_cols, collapse = ", "))

  if (!is.null(color_by)) {
    if (!is.character(color_by) || is.null(names(color_by)))
      stop("`color_by` must be a named character vector (names = sample IDs, values = group labels).")
    if (!color_by_id_col %in% names(x_df))
      stop("`color_by_id_col` '", color_by_id_col, "' not found in x_df.")
  }

  if (!"Cohort" %in% names(x_df))
    x_df <- mutate(x_df, Cohort = "All samples")

  # ── Helpers ──────────────────────────────────────────────────────────────────
  signif_label <- function(p) case_when(
    is.na(p)  ~ NA_character_,
    p <= 1e-4 ~ "****",
    p <= 1e-3 ~ "***",
    p <= 1e-2 ~ "**",
    p <= 5e-2 ~ "*",
    TRUE      ~ "ns"
  )

  fmt_p <- function(p)
    ifelse(is.na(p), NA_character_,
           ifelse(p < 0.001, "<0.001", format(signif(p, 2), trim = TRUE)))

  has_variation <- function(v) any(is.finite(v)) && sd(v, na.rm = TRUE) > 0

  safe_wilcox_p <- function(x, g) {
    g  <- droplevels(g)
    if (nlevels(g) != 2) return(NA_real_)
    x1 <- x[g == levels(g)[1]]; x2 <- x[g == levels(g)[2]]
    if (!length(x1) || !length(x2)) return(NA_real_)
    if (!any(is.finite(x1)) || !any(is.finite(x2))) return(NA_real_)
    if (sd(x1, na.rm = TRUE) == 0 && sd(x2, na.rm = TRUE) == 0 &&
        mean(x1, na.rm = TRUE) == mean(x2, na.rm = TRUE)) return(1)
    tryCatch(wilcox.test(x1, x2, exact = FALSE, conf.int = FALSE)$p.value,
             error = function(e) NA_real_)
  }

  percent_breaks_pretty5 <- function() {
    function(lims) {
      hi <- if (!is.finite(lims[2]) || lims[2] <= 0) 0.0004 else lims[2]
      px <- pretty(c(0, hi * 100), n = 4, min.n = 3)
      br <- sort(unique(c(0, px[px > 0]))) / 100
      if (length(br) > 5) {
        idx <- round(seq(1, length(br), length.out = 5))
        br  <- c(0, br[unique(idx[-1])])
      } else if (length(br) < 5) {
        nice_steps <- c(1e-6, 2e-6, 5e-6, 1e-5, 2e-5, 5e-5, 1e-4, 2e-4, 5e-4,
                        0.001, 0.002, 0.005, 0.01, 0.02, 0.05, 0.10, 0.20, 0.50, 1.00)
        idx  <- which(nice_steps * 4 >= hi)
        step <- if (length(idx)) nice_steps[idx[1]] else tail(nice_steps, 1)
        br   <- seq(0, step * 4, by = step)
      }
      br
    }
  }

  percent_labels_distinct <- function() {
    function(x) {
      ux   <- sort(unique(x[is.finite(x)]))
      step <- if (length(ux) >= 2) min(diff(ux)) else 0
      d    <- if (step > 0) max(0, min(4, ceiling(-log10(step * 100 + 1e-12)))) else 2
      fmt  <- function(v, k) paste0(formatC(v * 100, format = "f", digits = k), "%")
      lab  <- fmt(x, d)
      while (any(duplicated(lab)) && d < 4) { d <- d + 1; lab <- fmt(x, d) }
      lab
    }
  }

  y_pos_rel <- function(ymax, rank = 1L, rel_start = 0.12, rel_gap = 0.08) {
    rank <- as.integer(rank)
    add  <- rel_start + (rank - 1L) * rel_gap
    ifelse(is.finite(ymax) & ymax > 0,
           ymax * (1 + add),
           0.01 * (1 + (rank - 1L) * 0.8))
  }

  # ── Data prep ─────────────────────────────────────────────────────────────────
  df <- filter(x_df, Phenotype != "Other/Unassigned")

  if (!is.null(lineage_label))
    df <- filter(df, Lineage == lineage_label)

  df <- if (!is.null(cohort_levels)) {
    mutate(df, Cohort = factor(Cohort, levels = cohort_levels))
  } else {
    mutate(df, Cohort = forcats::fct_inorder(as.character(Cohort))) %>% 
      forcats::fct_na_level_to_value("All samples")
  }
  cohort_levels <- levels(df$Cohort)

  if (nrow(df) == 0)
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, size = 5,
                 label = paste("No data for plot; lineage:",
                               ifelse(is.null(lineage_label), "ALL", lineage_label))) +
        theme_void() + ggtitle("Phenotype proportions by cohort")
    )

  facet_levels <- if (is.list(phenotype_levels) && length(phenotype_levels) > 0) {
    if (!is.null(lineage_label) && lineage_label %in% names(phenotype_levels))
      phenotype_levels[[lineage_label]]
    else
      unlist(phenotype_levels, use.names = FALSE)
  } else {
    unique(as.character(df$Phenotype))
  }

  df <- mutate(df,
               Phenotype_chr = as.character(Phenotype),
               Phenotype_f   = factor(Phenotype_chr, levels = facet_levels, ordered = TRUE))

  # ── color_by join ─────────────────────────────────────────────────────────────
  use_color <- !is.null(color_by)

  if (use_color) {
    color_map <- tibble(
      !!color_by_id_col := names(color_by),
      .color_group       = unname(color_by)
    )
    df <- left_join(df, color_map, by = color_by_id_col)
  } else {
    df <- mutate(df, .color_group = NA_character_)
  }

  observed_cohort_lvls <- cohort_levels[
    cohort_levels %in% unique(as.character(df$Cohort[!is.na(df$Cohort)]))
  ]
  n_groups <- length(observed_cohort_lvls)
  can_test  <- isTRUE(add_tests) && n_groups %in% c(2, 3)

  # ── Panel maxima ──────────────────────────────────────────────────────────────
  peaks <- df %>%
    group_by(Phenotype_f) %>%
    summarise(ymax = max(prop_total, na.rm = TRUE), .groups = "drop") %>%
    mutate(ymax = ifelse(is.infinite(ymax), NA_real_, ymax))

  # ── Omnibus tests ─────────────────────────────────────────────────────────────
  strip_levels <- levels(df$Phenotype_f)
  omnibus      <- NULL

  if (can_test) {
    df_nonmiss <- filter(df, !is.na(prop_total))

    omnibus <- if (n_groups == 2) {
      df_nonmiss %>%
        group_by(Phenotype_f) %>%
        summarise(p = safe_wilcox_p(prop_total, Cohort), .groups = "drop") %>%
        mutate(Phenotype_f = as.character(Phenotype_f),
               p_txt = if_else(is.na(p), NA_character_,
                               paste0("Wilcoxon p = ", fmt_p(p))))
    } else {
      df_nonmiss %>%
        group_by(Phenotype_f) %>%
        filter(has_variation(prop_total)) %>%
        ungroup() %>%
        group_by(Phenotype_f) %>%
        rstatix::kruskal_test(prop_total ~ Cohort) %>%
        ungroup() %>%
        mutate(Phenotype_f = as.character(Phenotype_f),
               p_txt = paste0("Kruskal–Wallis p = ", fmt_p(p)))
    }

    p_text <- setNames(omnibus$p_txt, omnibus$Phenotype_f)

    strip_labels_expr <- setNames(
      ifelse(
        is.na(p_text[strip_levels]),
        paste0("bold('", gsub("'", "\\\\'", strip_levels, fixed = TRUE), "')"),
        paste0("atop(bold('", gsub("'", "\\\\'", strip_levels, fixed = TRUE),
               "'), '", gsub("'", "\\\\'", p_text[strip_levels], fixed = TRUE), "')")
      ),
      strip_levels
    )
  } else {
    strip_labels_expr <- setNames(
      paste0("bold('", gsub("'", "\\\\'", strip_levels, fixed = TRUE), "')"),
      strip_levels
    )
  }

  # ── Pairwise comparisons ──────────────────────────────────────────────────────
  pairwise_df <- NULL

  if (can_test) {
    df_nonmiss <- filter(df, !is.na(prop_total))

    if (n_groups == 2) {
      pairwise_df <- omnibus %>%
        filter(!is.na(p)) %>%
        transmute(
          Phenotype_f  = Phenotype_f,
          group1       = observed_cohort_lvls[1],
          group2       = observed_cohort_lvls[2],
          p.adj        = p,
          p.adj.signif = signif_label(p)
        ) %>%
        left_join(peaks, by = "Phenotype_f") %>%
        mutate(y.position = y_pos_rel(ymax, rank = 1L)) %>%
        filter(!is.na(p.adj.signif)) %>%
        select(-ymax)

    } else if (isTRUE(statisticallycorrect)) {
      eligible <- if (isTRUE(pairwise_if_omnibus_sig) && !is.null(omnibus)) {
        filter(omnibus, !is.na(p) & p < omnibus_alpha) %>% select(Phenotype_f)
      } else {
        distinct(df, Phenotype_f) %>% mutate(Phenotype_f = as.character(Phenotype_f))
      }

      keep3 <- df_nonmiss %>%
        semi_join(eligible, by = "Phenotype_f") %>%
        group_by(Phenotype_f) %>%
        summarise(k = n_distinct(Cohort), .groups = "drop") %>%
        filter(k == 3)

      if (nrow(keep3) > 0) {
        pairwise_df <- df_nonmiss %>%
          semi_join(keep3, by = "Phenotype_f") %>%
          group_by(Phenotype_f) %>%
          rstatix::dunn_test(prop_total ~ Cohort, p.adjust.method = "BH") %>%
          ungroup() %>%
          mutate(
            group1       = as.character(group1),
            group2       = as.character(group2),
            span         = abs(match(group2, cohort_levels) - match(group1, cohort_levels)),
            p.adj.signif = signif_label(p.adj)
          ) %>%
          group_by(Phenotype_f) %>%
          arrange(span, group1, group2, .by_group = TRUE) %>%
          mutate(rank = row_number()) %>%
          ungroup() %>%
          left_join(peaks, by = "Phenotype_f") %>%
          mutate(y.position = y_pos_rel(ymax, rank = rank)) %>%
          filter(!is.na(p.adj.signif)) %>%
          select(-span, -rank, -ymax)
      }

    } else {
      ref_level <- observed_cohort_lvls[1]
      others    <- setdiff(observed_cohort_lvls, ref_level)

      pairwise_df <- expand.grid(
        Phenotype_f = levels(df$Phenotype_f),
        group2      = others,
        stringsAsFactors = FALSE
      ) %>%
        mutate(group1 = ref_level) %>%
        rowwise() %>%
        mutate(p = {
          sub <- df_nonmiss %>%
            filter(as.character(Phenotype_f) == .data$Phenotype_f,
                   as.character(Cohort) %in% c(group1, group2))
          if (n_distinct(sub$Cohort) < 2) 1 else safe_wilcox_p(sub$prop_total, droplevels(sub$Cohort))
        }) %>%
        ungroup() %>%
        group_by(Phenotype_f) %>%
        mutate(
          p.adj        = p.adjust(p, method = "BH"),
          p.adj.signif = signif_label(p.adj),
          rank         = match(group2, others)
        ) %>%
        ungroup() %>%
        left_join(peaks, by = "Phenotype_f") %>%
        mutate(y.position = y_pos_rel(ymax, rank = rank)) %>%
        filter(!is.na(p.adj.signif)) %>%
        select(-rank, -ymax)
    }
  }

  top_expand <- if (!is.null(pairwise_df) && nrow(pairwise_df) > 0) 0.18 else 0.06

  # ── Plot ──────────────────────────────────────────────────────────────────────
  p <- ggplot(df, aes(x = Cohort, y = prop_total)) +
    geom_boxplot(
      width        = box_width,
      outlier.shape = NA,
      linewidth    = 0.6,
      fill         = "grey90"
    ) +
    geom_point(
      aes(color = if (use_color) .color_group else NULL),
      position = position_jitter(width = 0.08, height = 0),
      size     = point_size,
      alpha    = point_alpha
    ) +
    facet_wrap(
      ~ Phenotype_f,
      ncol      = ncol,
      scales    = "free_y",
      drop      = FALSE,
      labeller  = as_labeller(strip_labels_expr, label_parsed)
    ) +
    scale_y_continuous(
      breaks       = percent_breaks_pretty5(),
      labels       = percent_labels_distinct(),
      minor_breaks = NULL,
      limits       = function(lims) c(0, lims[2]),
      expand       = expansion(mult = c(0, top_expand))
    ) +
    labs(
      x        = if (n_groups == 1) NULL else "Group",
      y        = "Percent of Total Cells",
      subtitle = subtitle
    ) +
    theme_base +
    theme(
      strip.placement  = "outside",
      strip.text       = element_text(size = 8.5, lineheight = 1.05,
                                      margin = margin(t = 4, r = 6, b = 4, l = 6)),
      strip.background = element_rect(fill = "white", color = "grey60", linewidth = 0.3),
      axis.title.x     = element_text(size = 9),
      axis.title.y     = element_text(size = 9),
      axis.text.x      = if (n_groups == 1) element_blank() else element_text(size = 9, face = "bold"),
      axis.ticks.x     = if (n_groups == 1) element_blank() else element_line(),
      axis.text.y      = element_text(size = 8),
      panel.grid.major = element_line(color = "grey90", linewidth = 0.25),
      panel.grid.minor = element_blank(),
      panel.spacing.x  = grid::unit(12, "pt"),
      panel.spacing.y  = grid::unit(27, "pt"),
      plot.margin      = margin(t = 16, r = 10, b = 8, l = 8)
    ) +
    coord_cartesian(clip = "off")

  if (use_color)
    p <- p +
      scale_color_brewer(palette = "Set1", na.value = "grey50", name = NULL) +
      theme(legend.position = "bottom")

  if (!is.null(pairwise_df) && nrow(pairwise_df) > 0)
    p <- p + ggpubr::stat_pvalue_manual(
      data         = pairwise_df,
      label        = "p.adj.signif",
      xmin         = "group1",
      xmax         = "group2",
      y.position   = "y.position",
      tip.length   = 0.01,
      bracket.size = 0.35,
      hide.ns      = !isTRUE(show_ns),
      size         = 2.7
    )

  p
}

```

## Background and Objectives

## Methods

```{r gating summary}

spatial_df_final %>%
  filter(!Phenotype %in% c("Other/Unassigned", "Ambiguous/Double+")) %>%
  count(Lineage, Phenotype) %>%
  group_by(Lineage) %>%
  mutate(`% of lineage` = round(n / sum(n) * 100, 2)) %>%
  ungroup() %>%
  arrange(Lineage, desc(n)) %>%
  select(Lineage, Phenotype) %>%
  gt::gt(groupname_col = "Lineage") %>%
  gt::tab_caption("Phenotype definitions by lineage") %>%
  gt::cols_width(
    Lineage  ~ gt::px(100),
    Phenotype ~ gt::px(350)
  ) %>%
  gt::cols_align(align = "left") %>%
  gt::tab_options(
    table.width          = gt::px(460),
    table.align          = "center",
    data_row.padding     = gt::px(3),
    table.font.size      = gt::px(13),
    row_group.font.weight = "bold"
  )

```

## Results

### Abundance

```{r Table1}

make_phenotype_tbl_summary <- function(
  df,
  cohort_levels     = NULL,
  lineage_levels    = NULL,
  phenotype_levels  = NULL,
  label_px          = 260,
  column_px         = 120,
  font_size_px      = 13,
  row_padding_px    = 3,
  summary_padding_px= 2
) {

  # ----------------------------- Validation -----------------------------------
  req <- c("SampleID", "Lineage", "Phenotype", "prop_total")
  missing_cols <- setdiff(req, names(df))

  if (length(missing_cols) > 0) {
    stop("`df` must contain: ", paste(missing_cols, collapse = ", "))
  }

  has_cohort <- "Cohort" %in% names(df)

  if (has_cohort) {
    df <- df %>%
      mutate(Cohort = as.character(.data$Cohort))
  
    observed_cohorts <- unique(df$Cohort[!is.na(df$Cohort) & nzchar(df$Cohort)])
  
    if (!is.null(cohort_levels) && length(cohort_levels) > 0) {
      cohort_levels <- intersect(cohort_levels, observed_cohorts)
    } else {
      cohort_levels <- observed_cohorts
    }
  
    if (length(cohort_levels) > 0) {
      df <- df %>%
        filter(.data$Cohort %in% cohort_levels) %>%
        mutate(Cohort = factor(.data$Cohort, levels = cohort_levels))
    } else {
      has_cohort <- FALSE
      df <- df %>% select(-.data$Cohort)
    }
  }

  # ----------------------------- Phenotype order ------------------------------
  present <- df %>%
    distinct(.data$Lineage, .data$Phenotype)

  if (is.list(phenotype_levels) && length(phenotype_levels) > 0) {
    lineage_order <- if (!is.null(lineage_levels)) {
      lineage_levels
    } else {
      names(phenotype_levels)
    }

    phen_order <- phenotype_levels[
      intersect(lineage_order, names(phenotype_levels))
    ] %>%
      unlist(use.names = FALSE) %>%
      intersect(present$Phenotype)
  } else {
    if (!is.null(lineage_levels)) {
      present <- present %>%
        mutate(Lineage = factor(.data$Lineage, levels = lineage_levels))
    }

    phen_order <- present %>%
      arrange(.data$Lineage, .data$Phenotype) %>%
      pull(.data$Phenotype)
  }

  # ----------------------------- Wide table -----------------------------------
  id_cols <- if (has_cohort) c("SampleID", "Cohort") else "SampleID"

  wide_df <- df %>%
    select(all_of(id_cols), .data$Phenotype, .data$prop_total) %>%
    group_by(across(all_of(c(id_cols, "Phenotype")))) %>%
    summarise(
      prop_total = mean(.data$prop_total, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from = .data$Phenotype,
      values_from = .data$prop_total
    ) %>%
    mutate(across(-all_of(id_cols), ~ as.numeric(.x * 100)))

  selected_phenotypes <- intersect(phen_order, setdiff(names(wide_df), id_cols))

  wide_df <- wide_df %>%
    select(any_of("Cohort"), all_of(selected_phenotypes))

  # ----------------------------- Summary table --------------------------------
  tbl_args <- list(
    data      = wide_df,
    include   = where(is.numeric),
    type      = everything() ~ "continuous2",
    missing   = "no",
    statistic = all_continuous() ~ c(
      "{mean} ({sd})",
      "{median} [{min}, {max}]"
    ),
    digits = all_continuous() ~ 2
  )

  if (has_cohort) {
    tbl_args$by <- quote(Cohort)
  }

  tbl <- do.call(gtsummary::tbl_summary, tbl_args) %>%
    gtsummary::modify_table_body(function(tb) {
      stat_cols <- grep("^stat_", names(tb), value = TRUE)

      stat_mat <- if (length(stat_cols) > 0) {
        as.data.frame(tb[stat_cols], stringsAsFactors = FALSE)
      } else {
        NULL
      }

      has_stat <- if (!is.null(stat_mat)) {
        apply(stat_mat, 1, function(r) any(nzchar(r) & !is.na(r)))
      } else {
        rep(FALSE, nrow(tb))
      }

      is_median <- if (!is.null(stat_mat)) {
        apply(stat_mat, 1, function(r) any(grepl("\\[", r)))
      } else {
        rep(FALSE, nrow(tb))
      }

      is_mean <- has_stat & !is_median

      if ("row_type" %in% names(tb) && "label" %in% names(tb)) {
        ix <- tb$row_type == "label" & grepl("\\([^)]*\\+[^)]*\\)", tb$label)

        tb$label[ix] <- sub("\\s*\\(", "<br/>(", tb$label[ix])

        tb$label[is_mean]   <- "Mean (SD)"
        tb$label[is_median] <- "Median [Min, Max]"
      }

      if (length(stat_cols) > 0) {
        for (sc in stat_cols) {
          z <- tb[[sc]]

          z[!is.na(z) & z == "0.00 (0.00)"] <- "<0.01"

          idx <- !is.na(z) & is_mean & grepl("^0\\.00 \\(", z)
          if (any(idx)) {
            z[idx] <- sub("^0\\.00", "<0.01", z[idx])
          }

          idx <- !is.na(z) & is_median
          if (any(idx)) {
            z[idx] <- gsub("\\b0\\.00\\b", "<0.01", z[idx])
          }

          tb[[sc]] <- z
        }
      }

      tb
    }) %>%
    gtsummary::modify_header(label ~ "**Phenotype**")

  if (has_cohort) {
    tbl <- tbl %>%
      gtsummary::modify_header(gtsummary::all_stat_cols() ~ "**{level} (%)**") %>%
      gtsummary::modify_caption(
        "**Table 1:** Summary of mean, standard deviation, median, and range of percent abundance for each immune phenotype across sample cohorts. Values represent cell-type frequencies as a percentage of all segmented cells within each sample."
      )
  } else {
    tbl <- tbl %>%
      gtsummary::modify_header(gtsummary::all_stat_cols() ~ "**Overall (%)**") %>%
      gtsummary::modify_caption(
        "**Table 1:** Summary of mean, standard deviation, median, and range of percent abundance for each immune phenotype across samples. Values represent cell-type frequencies as a percentage of all segmented cells within each sample."
      )
  }

  tbl <- tbl %>%
    gtsummary::bold_labels()

  # ----------------------------- gt formatting --------------------------------
  gt_tbl <- gtsummary::as_gt(tbl)

  bh <- gt_tbl[["_boxhead"]]

  gt_cols <- if (!is.null(bh) && "var" %in% names(bh)) {
    unique(bh$var[!is.na(bh$var)])
  } else {
    names(gt_tbl[["_data"]])
  }

  label_col <- intersect("label", gt_cols)
  num_cols  <- setdiff(gt_cols, "label")

  width_fmls <- list()

  if (length(label_col) == 1) {
    width_fmls <- c(
      width_fmls,
      rlang::new_formula(
        rlang::sym(label_col),
        gt::px(as.integer(label_px))
      )
    )
  }

  if (length(num_cols) > 0) {
    width_fmls <- c(
      width_fmls,
      lapply(
        num_cols,
        function(col) {
          rlang::new_formula(
            rlang::sym(col),
            gt::px(as.integer(column_px))
          )
        }
      )
    )
  }

  if (length(width_fmls) > 0) {
    gt_tbl <- do.call(gt::cols_width, c(list(gt_tbl), width_fmls))
  }

  if (length(num_cols) > 0) {
    gt_tbl <- gt::cols_align(gt_tbl, columns = num_cols, align = "center")
  }

  if (length(label_col) == 1) {
    gt_tbl <- gt_tbl %>%
      gt::cols_align(columns = label_col, align = "left") %>%
      gt::fmt_markdown(columns = label_col)
  }

  gt_tbl %>%
    gt::tab_options(
      data_row.padding          = gt::px(as.integer(row_padding_px)),
      summary_row.padding       = gt::px(as.integer(summary_padding_px)),
      grand_summary_row.padding = gt::px(as.integer(summary_padding_px)),
      table.font.size           = gt::px(as.integer(font_size_px))
    )
}

make_phenotype_tbl_summary(
  df = abundance_plot_df,
  phenotype_levels = phenotype_levels, 
  cohort_levels = NULL,
  label_px = 300,
  column_px = 120,
  font_size_px = 14,
  row_padding_px = 3,
  summary_padding_px = 2
)

```

```{r Boxplots}
#| results: asis
#| echo: false
#| message: false
#| warning: false
#| fig-width: 11       # bigger intrinsic figure size (inches)
#| fig-height: 6
#| out-width: "100%"   # scale to the tab’s full width
#| fig-align: center   # center the plot
#| dpi: 200            # crisper PNGs
#| fig-retina: 2       # high-DPI displays

# Specify color labels for different SampleID with named vector  (optional)

plots <- setNames(
  map(names(phenotype_levels), function(lin)
    plot_lineage(abundance_plot_df, lin,
                 phenotype_levels  = phenotype_levels,
                 cohort_levels     = cohort_levels,
                 ncol              = 3,
                 add_tests         = FALSE,
                 statisticallycorrect = FALSE,
                 subtitle          = NULL,
                 color_by          = setNames(metadata$Class, metadata$SampleID),
                 color_by_id_col   = "SampleID")
  ),
  names(phenotype_levels)
)

cat("::: {.panel-tabset}\n\n")  # open the tabset

walk(seq_along(plots), function(i) {
  nm <- names(plots)[i]
  cat("### ", nm, "\n\n", sep = "")  # tab heading
  print(plots[[i]])
  cat("\n\n")  # important: two newlines to separate tabs
})

cat(":::\n")  # close the tabset

```

**Figure 1:** Boxplots of percent abundance for each immune cell phenotype out of all cells in the sample. Colored dots indicate primary (blue) or metastatic (red) tumors.

```{r phenotypeproptable}
#| fig-width: 11       # bigger intrinsic figure size (inches)
#| fig-height: 6
#| out-width: "100%"   # scale to the tab’s full width
#| 

spatial_df_final %>%
  filter(!Phenotype %in% c("Other/Unassigned", "Ambiguous/Double+")) %>%
  count(Lineage, Phenotype) %>%
  group_by(Lineage) %>%
  mutate(`% of lineage` = round(n / sum(n) * 100, 2)) %>%
  ungroup() %>%
  arrange(Lineage, desc(n)) %>%
  kbl(caption = "Phenotype composition within each lineage (excluding Other/Unassigned and Ambiguous)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = T) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  scroll_box(width = "100%")

```

```{=html}
<div style="width:100%;"></div>
```
```{r stackedbars}
#| fig-width: 11       # bigger intrinsic figure size (inches)
#| fig-height: 6
#| out-width: "100%"   # scale to the tab’s full width
#| 

.sid_order <- sort(unique(abundance_plot_df$SampleID))

.lineage_levels <- names(phenotype_levels)

.pheno_names <- unlist(phenotype_levels, use.names = F)

.pheno_palette <- setNames(
  c(
    colorRampPalette(
      c(
        "#E31A1C",  # red
        "#1F78B4",  # blue
        "#33A02C",  # green
        "#FF7F00",  # orange
        "#6A3D9A",  # purple
        "#B15928",  # brown
        "#00CED1",  # dark turquoise
        "#FF1493",  # deep pink
        "#FFD700",  # gold
        "#000080",  # navy
        "#7FFF00",  # chartreuse
        "#DC143C",  # crimson
        "#00FA9A",  # medium spring green
        "#FF6347",  # tomato
        "#4B0082",  # indigo
        "#F0E68C",  # khaki
        "#8B0000",  # dark red
        "#00BFFF"   # deep sky blue
      )
    )(length(.pheno_names)),
    "grey80"
  ),
  c(.pheno_names, "Other/Unassigned")
)

p1 <- abundance_plot_df %>%
  filter(!Lineage %in% c("Other/Unassigned", "Ambiguous/Double+")) %>%
  mutate(
    SampleID = factor(SampleID, levels = .sid_order),
    Lineage  = factor(as.character(Lineage), levels = .lineage_levels)
  ) %>%
  group_by(SampleID, Lineage) %>%
  summarise(pct = sum(prop_total) * 100, .groups = "drop") %>%
  ggplot(aes(x = SampleID, y = pct, fill = Lineage)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.3) +
  scale_fill_brewer(palette = "Set2", name = "Lineage") +
  scale_x_discrete(limits = .sid_order) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(title = "Lineage", x = NULL, y = "% of total cells") +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x        = element_text(angle = 45, hjust = 1, size = 8),
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", hjust = 0.5),
    legend.position    = "bottom",
    legend.title       = element_text(face = "bold", size = 9),
    legend.text        = element_text(size = 10),
    legend.key.size    = unit(9, "pt")
  )

p2 <- abundance_plot_df %>%
  filter(!Phenotype %in% c("Other/Unassigned", "Ambiguous/Double+")) %>%
  mutate(
    SampleID  = factor(SampleID, levels = .sid_order),
    Phenotype = factor(as.character(Phenotype), levels = .pheno_names)
  ) %>%
  group_by(SampleID, Phenotype) %>%
  summarise(pct = sum(prop_total) * 100, .groups = "drop") %>%
  ggplot(aes(x = SampleID, y = pct, fill = Phenotype)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.15) +
  scale_fill_manual(values = .pheno_palette, name = "Phenotype") +
  scale_x_discrete(limits = .sid_order) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(title = "Phenotype", x = NULL, y = NULL) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x        = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title         = element_text(face = "bold", hjust = 0.5),
    legend.position    = "bottom",
    legend.title       = element_text(face = "bold", size = 9),
    legend.text        = element_text(size = 8.5),
    legend.key.size    = unit(8, "pt")
  ) +
  guides(fill = guide_legend(nrow = 7, ncol = 2, byrow = TRUE))

p1 + p2 +
  plot_layout(widths = c(1, 1.8), nrow = 1)

```

**Figure 2:** Stacked bar charts of observed immune cell abundance (%) in relation to the total cell population. Abundances are organized by lineage and individual phenotypes within the lineage.

```{r marker-hierarchy, results='asis', dpi=96}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 11       # bigger intrinsic figure size (inches)
#| fig-height: 6
#| out-width: "100%"   # scale to the tab’s full width
#| 
.marker_matrix <- spatial_df_final %>%
  select(all_of(cfg$gammagater$markers)) %>%
  mutate(across(everything(), as.integer)) %>%
  as.matrix()

cat("::: {.panel-tabset}\n\n")

for (focal_marker in cfg$gammagater$markers) {

  cat("##", focal_marker, "\n\n")

  other_markers <- setdiff(cfg$gammagater$markers, focal_marker)
  focal_rows    <- .marker_matrix[, focal_marker] == 1L
  n_focal       <- sum(focal_rows)

  if (n_focal < 10) {
    cat("Insufficient", focal_marker, "+ cells for analysis.\n\n")
    next
  }

  # ── Build focal matrix ──────────────────────────────────────────────────────
  focal_mat <- matrix(
    as.integer(.marker_matrix[focal_rows, other_markers]),
    nrow = n_focal
  )
  colnames(focal_mat) <- other_markers

  # ── Compute intersection counts via bit keys ────────────────────────────────
  powers      <- 2L ^ (seq_len(length(other_markers)) - 1L)
  keys        <- as.integer(focal_mat %*% powers)
  key_counts  <- table(keys)
  unique_keys <- as.integer(names(key_counts))

  unique_mat           <- outer(unique_keys, powers, FUN = bitwAnd) > 0L
  colnames(unique_mat) <- other_markers

  combo_strings <- apply(unique_mat, 1, function(row) {
    pos <- other_markers[as.logical(row)]
    if (length(pos) == 0L) "None" else paste(sort(pos), collapse = " + ")
  })

  combo_df <- data.frame(
    combo = combo_strings[match(as.character(keys),
                                as.character(unique_keys))],
    stringsAsFactors = FALSE
  ) %>%
    dplyr::count(combo) %>%
    dplyr::mutate(
      pct       = round(n / sum(n) * 100, 2),
      n_markers = if_else(
        combo == "None", 0L,
        as.integer(lengths(strsplit(combo, " + ", fixed = TRUE)))
      )
    ) %>%
    # ── Aggregate rare combinations below 0.4% into "Other combinations" ─────
    dplyr::mutate(
      combo     = if_else(pct < 0.4, "Other combinations", combo),
      n_markers = if_else(pct < 0.4, NA_integer_, n_markers)
    ) %>%
    dplyr::group_by(combo, n_markers) %>%
    dplyr::summarise(
      n   = sum(n),
      pct = sum(pct),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      label = paste0(combo, "\n", round(pct, 2), "%")
    ) %>%
    dplyr::arrange(desc(n))

  # ── Treemap ──────────────────────────────────────────────────────────────────
  p_treemap <- ggplot(combo_df,
                      aes(area = n, fill = n_markers, label = label)) +
    geom_treemap(colour = "black", linewidth = 0.8) +
    geom_treemap_text(
      color     = "black",
      place     = "centre",
      size      = 9,
      fontface  = "bold",
      reflow    = T,
      padding.x = grid::unit(2, "mm"),
      padding.y = grid::unit(2, "mm")
    ) +
    scale_fill_distiller(
      palette  = "Blues",
      direction = 1,
      name     = "N co-markers",
      na.value = "grey80"
    ) +
    labs(
      title    = paste0("Co-marker treemap among ", focal_marker, "+ cells"),
      subtitle = paste0("N = ", format(n_focal, big.mark = ","),
                        " ", focal_marker, "+ cells. ",
                        "Combinations <0.4% aggregated into 'Other combinations'.")
    ) +
    theme_bw(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))

  # ── UpSet ─────────────────────────────────────────────────────────────────── 
  p_upset <- tryCatch({
    upset_df <- as.data.frame(focal_mat == 1L)

    if (nrow(upset_df) > 50000) {
      set.seed(42)
      upset_df <- upset_df[sample(nrow(upset_df), 50000), , drop = FALSE]
    }

    upset(
      upset_df,
      intersect          = other_markers,
      base_annotations   = list(
        "Intersection size" = intersection_size(counts = TRUE)
      ),
      width_ratio        = 0.2,
      min_size           = max(1, floor(nrow(upset_df) * 0.005)),
      sort_sets          = "descending",
      sort_intersections = "descending"
    ) +
      labs(
        title    = paste0("Co-marker positivity among ",
                          focal_marker, "+ cells"),
        subtitle = paste0(
          "N = ", format(n_focal, big.mark = ","), " total ",
          focal_marker, "+ cells; UpSet shown on 50,000-cell subsample"
        )
      ) +
      theme_bw(base_size = 11)
  }, error = function(e) NULL)

  if (!is.null(p_upset))   { print(p_upset);   cat("\n\n") }
  if (!is.null(p_treemap)) { print(p_treemap); cat("\n\n") }
  
}

cat(":::\n")

```

**Figure 3:** UpSet plot and treemap of most prevalent phenotypes when a cell is positive for a given marker. Phenotypes with \<0.4% prevalence are aggregated into an "Other combinations" category.

### Spatial Visualization

```{r}
#| results: asis
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 6
#| dpi: 400
#| fig-align: center

plot_spatial_tabs_html <- function(
  df,
  out_dir           = "spatial_tab_imgs",
  fg_point_size     = 0.70,
  bg_point_size     = 0.25,
  alpha_fg          = 0.60,
  alpha_bg          = 0.35,
  flip_y            = FALSE,
  legend_text_size  = 14,
  legend_wrap_width = 16,
  legend_max_rows   = 3,
  legend_dot_size   = 5,
  fig_width         = 20,
  fig_height        = 14,
  dpi               = 200,
  group_col         = NULL,
  color_palette     = NULL
) {

  # ── Validation ───────────────────────────────────────────────────────────────
  stopifnot(all(c("SampleID", "X", "Y", "Phenotype") %in% names(df)))
  df$X <- as.numeric(df$X)
  df$Y <- as.numeric(df$Y)

  other_name <- "Other/Unassigned"
  df$Phenotype[df$Phenotype == "" | is.na(df$Phenotype)] <- other_name

  # ── Helpers ──────────────────────────────────────────────────────────────────
  wrap_labels <- function(x, width)
    vapply(x, function(s)
      paste(strwrap(as.character(s), width = width), collapse = "\n"),
      "", USE.NAMES = FALSE)

  idify <- function(x) gsub("[^A-Za-z0-9_-]+", "-", x)
  fwdsl <- function(p) gsub("\\\\", "/", p)

  # ── Output directory ─────────────────────────────────────────────────────────
  out_dir <- fwdsl(normalizePath(out_dir, mustWork = FALSE))
  if (!dir.exists(out_dir))
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  # ── Global levels and palette ────────────────────────────────────────────────
  rest_global <- sort(unique(df$Phenotype[df$Phenotype != other_name]))
  global_levs <- c(other_name, rest_global)

  if (!is.null(color_palette)) {
    # Use supplied palette — reorder to match global_levs, fill any gaps
    cols_global <- color_palette
    cols_global[other_name] <- "grey80"
    global_levs <- c(
      other_name,
      names(color_palette)[names(color_palette) != "other_name"]
    )
    global_levs <- unique(global_levs)
  } else {
    rest_global <- sort(unique(df$Phenotype[df$Phenotype != other_name]))
    global_levs <- c(other_name, rest_global)
  }

  # ── Tab skeleton ─────────────────────────────────────────────────────────────
  sids       <- sort(unique(df$SampleID))
  tab_prefix <- paste0("sv-", as.integer(as.numeric(Sys.time())))

  cat(sprintf(
    '<ul class="nav nav-tabs" id="%s-tablist" role="tablist">\n',
    tab_prefix
  ))
  for (i in seq_along(sids)) {
    sid        <- sids[i]
    tab_id     <- paste0(tab_prefix, "-tab-",  i)
    pane_id    <- paste0(tab_prefix, "-pane-", i)
    active_cls <- if (i == 1) " active" else ""
    aria_sel   <- if (i == 1) "true" else "false"
    cat(sprintf(
      '<li class="nav-item" role="presentation"><button class="nav-link%s" id="%s" data-bs-toggle="tab" data-bs-target="#%s" type="button" role="tab" aria-controls="%s" aria-selected="%s">%s</button></li>\n',
      active_cls, tab_id, pane_id, pane_id, aria_sel, sid
    ))
  }
  cat("</ul>\n")

  # ── Render each sample ───────────────────────────────────────────────────────
  cat(sprintf(
    '<div class="tab-content" id="%s-content" style="margin-top:10px;">\n',
    tab_prefix
  ))

  for (i in seq_along(sids)) {
    sid         <- sids[i]
    pane_id     <- paste0(tab_prefix, "-pane-", i)
    tab_id      <- paste0(tab_prefix, "-tab-",  i)
    show_active <- if (i == 1) " show active" else ""

    d           <- df[df$SampleID == sid, , drop = FALSE]
    d$Phenotype <- factor(d$Phenotype, levels = global_levs)

    present     <- intersect(global_levs, unique(as.character(d$Phenotype)))
    lab_wrapped <- wrap_labels(present, legend_wrap_width)

    rows <- max(1, min(legend_max_rows,
                       ceiling(length(present) /
                               max(1, floor(fig_width /
                                            (legend_wrap_width * 0.6))))))

    d_bg <- d[as.character(d$Phenotype) == other_name, , drop = FALSE]
    d_fg <- d[as.character(d$Phenotype) != other_name, , drop = FALSE]

    # Title (SampleID & cell count)
    title_label <- {
      n_cells <- sum(is.finite(d$X) & is.finite(d$Y))
      if (!is.null(group_col) && group_col %in% names(d)) {
        grp <- unique(d[[group_col]])
        grp <- grp[!is.na(grp)]
        paste0(sid, " (", paste(grp, collapse = "/"), ") | N = ",
               format(n_cells, big.mark = ","))
      } else {
        paste0(sid, " - N = ", format(n_cells, big.mark = ","))
      }
    }

    p <- ggplot() +
      { if (nrow(d_bg) > 0)
          geom_scattermore(data = d_bg,
                     aes(X, Y, color = Phenotype),
                     alpha = alpha_bg, size = bg_point_size, shape = 16) } +
      { if (nrow(d_fg) > 0)
          geom_scattermore(data = d_fg,
                     aes(X, Y, color = Phenotype),
                     alpha = alpha_fg, size = fg_point_size, shape = 16) } +
      coord_equal() +
      { if (flip_y) scale_y_reverse() } +
      scale_color_manual(
        values = cols_global,
        breaks = present,
        labels = lab_wrapped,
        drop   = FALSE
      ) +
      labs(title = title_label) +
      theme_void(base_size = 13) +
      theme(
        plot.title       = element_text(size = 14, face = "bold",
                                        hjust = 0.5,
                                        margin = margin(b = 8)),
        legend.position  = "bottom",
        legend.direction = "horizontal",
        legend.box       = "horizontal",
        legend.text      = element_text(size = legend_text_size),
        legend.title     = element_blank(),
        legend.margin    = margin(2, 2, 2, 2),
        legend.spacing.y = unit(6, "pt"),
        legend.spacing.x = unit(8, "pt")
      ) +
      guides(color = guide_legend(
        title        = NULL,
        nrow         = rows,
        byrow        = TRUE,
        override.aes = list(size = legend_dot_size, alpha = 1)
      ))

    fname <- paste0("spatial_", idify(sid), ".png")
    fpath <- fwdsl(file.path(out_dir, fname))
    ggsave(fpath, plot = p,
           width = fig_width, height = fig_height,
           dpi = dpi, limitsize = FALSE)

    img_tag <- sprintf(
      '<img src="%s" style="width:100%%; height:80vh; object-fit:contain; display:block; margin:0 auto;" />',
      fpath
    )
    cat(sprintf(
      '<div class="tab-pane fade%s" id="%s" role="tabpanel" aria-labelledby="%s">\n  %s\n</div>\n',
      show_active, pane_id, tab_id, img_tag
    ))
  }

  cat("</div>\n")
  invisible(NULL)
}

```

```{r spatial}
#| results: asis
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
#| out-width: "100%"
#| fig-align: center

plot_spatial_tabs_html(
  df = spatial_df_final %>% filter(Phenotype != "Ambiguous/Double+"),
  flip_y = TRUE,
  fg_point_size = 1.4, # Lower if dots forming pixel structure or obscuring each other, raise if lots of cells 
  bg_point_size = 0.2,
  alpha_fg = 0.80,
  alpha_bg = 0.2,        
  legend_text_size = 12,   
  legend_wrap_width = 9,
  legend_max_rows = 3,
  fig_width = 10,
  fig_height = 8,
  dpi = 300,
  color_palette = .pheno_palette
)

```

## Conclusions

### Algorithm Parameters

To ensure reproducibility of the processing workflow, click "Show the code" to view the scripts and/or their corresponding parametrizations for obtaining segmentation and phenotyping results.

```{r}

# Segmentation: Stardist

# def pLow = 2; def pHigh = 98   
# def threshold = 0.6            
# def pixelSize = 0.7            
# def cellExpand = 3             

# Phenotyping: GammaGateR

# boundaries_IO60 <- list(
#   `b-Catenin1`        = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `Bcl-2`             = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `Beta-actin`        = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   Caveolin            = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD107a              = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD11b               = matrix(c(0, 0.45, 0.50, 1), byrow = TRUE, nrow = 2),
#   CD11c               = matrix(c(0, 0.25, 0.35, Inf), byrow = TRUE, nrow = 2),
#   CD14                = matrix(c(0, 0.2, 0.30, 1), byrow = TRUE, nrow = 2),
#   CD163               = matrix(c(0, 0.275, 0.3, Inf), byrow = TRUE, nrow = 2),
#   CD20                = matrix(c(0, 0.55, 0.675, Inf), byrow = TRUE, nrow = 2),
#   CD206               = matrix(c(0, 0.275, 0.4, Inf), byrow = TRUE, nrow = 2),
#   CD209               = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD21                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD31                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD34                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD38                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD39                = matrix(c(0, 0.075, 0.1, Inf), byrow = TRUE, nrow = 2),
#   CD3e                = matrix(c(0, 0.3, 0.4, Inf), byrow = TRUE, nrow = 2),
#   CD4                 = matrix(c(0, 0.35, 0.4, Inf), byrow = TRUE, nrow = 2),
#   CD44                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD45                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD45RO              = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD56                = matrix(c(0, 0.325, 0.45, Inf), byrow = TRUE, nrow = 2),
#   CD57                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD66                = matrix(c(0, 0.35, 0.45, Inf), byrow = TRUE, nrow = 2),
#   CD68                = matrix(c(0, 0.2, 0.25, Inf), byrow = TRUE, nrow = 2),
#   CD79a               = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   CD8                 = matrix(c(0, 0.275, 0.3, Inf), byrow = TRUE, nrow = 2),
#   `Collagen IV`       = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   DAPI                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `E-cadherin`        = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   EpCAM               = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   ER                  = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   FOXP3               = matrix(c(0, 0.3, 0.40, Inf), byrow = TRUE, nrow = 2),
#   GP100               = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `Granzyme B`        = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `Histone H3 (p Ser28)` = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `HLA-A`             = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `HLA-DR`            = matrix(c(0, 0.25, 0.30, Inf), byrow = TRUE, nrow = 2),
#   ICOS                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   IDO1                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   IFNG                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   iNOS                = matrix(c(0, 0.3, 0.375, Inf), byrow = TRUE, nrow = 2),
#   `Keratin 14`        = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `Keratin 5`         = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `Keratin 8/18`      = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   Ki67                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   LAG3                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   MPO                 = matrix(c(0, 0.25, 0.325, Inf), byrow = TRUE, nrow = 2),
#   `Pan-Cytokeratin`   = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   PCNA                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `PD-1`              = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `PD-L1`             = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   Podoplanin          = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   SMA                 = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   SOX2                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   `TCF-1`             = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   TOX                 = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   TP63                = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   Vimentin            = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2),
#   VISTA               = matrix(c(0, 0.25, 0.50, Inf), byrow = TRUE, nrow = 2)
# )

# Phenotyping: FDR Classification

# marker_fdr <- c(
#   CD3e     = 0.10, CD4   = 0.20, CD8   = 0.20, FOXP3 = 0.20, CD39  = 0.20,
#   CD56     = 0.05, CD20  = 0.15,
#   CD66     = 0.10, MPO   = 0.10,
#   CD68     = 0.05, CD163 = 0.10, CD206 = 0.10, iNOS  = 0.10,
#   CD11b    = 0.10, CD14  = 0.10, CD11c = 0.10,
#   `HLA-DR` = 0.10
# )
# 
# marker_min_pos <- setNames(rep(0.001, length(marker_fdr)), names(marker_fdr))
# 
# marker_max_cap <- c(
#   CD3e     = 0.05, CD4   = 0.05, CD8   = 0.05, FOXP3 = 0.05, CD39  = 0.05,
#   CD56     = 0.05, CD20  = 0.08,
#   CD66     = 0.15, MPO   = 0.15,
#   CD68     = 0.20, CD163 = 0.25, CD206 = 0.25, iNOS  = 0.20,
#   CD11b    = 0.40, CD14  = 0.30, CD11c = 0.20,
#   `HLA-DR` = 0.50
# )

```

### Session Info

```{r}

devtools::session_info()

```

### References

1.  Akoya Biosciences, Inc. (2023). *PhenoCycler‑Fusion user guide* (Doc. No. PD‑000011, Rev. K). https://content.protocols.io/files/nur5b8i4p.pdf

2.  Bankhead, P., Loughrey, M. B., Fernández, J. A., Dombrowski, Y., McArt, D. G., Dunne, P. D., McQuaid, S., Gray, R. T., Murray, L. J., Coleman, H. G., James, J. A., Salto-Tellez, M., & Hamilton, P. W. (2017). QuPath: Open source software for digital pathology image analysis. *Scientific Reports*, *7*(1). https://doi.org/10.1038/s41598-017-17204-5

3.  Schmidt, U., Weigert, M., Broaddus, C., & Myers, G. (2018). Cell Detection with Star-convex Polygons. *ArXiv:1806.03535 \[Cs\]*, *11071*, 265–273. https://doi.org/10.1007/978-3-030-00934-2_30

4.  Xiong, J., Kaur, H., Heiser, C. N., McKinley, E. T., Roland, J. T., Coffey, R. J., Shrubsole, M. J., Wrobel, J., Ma, S., Lau, K. S., & Vandekar, S. (2024). GammaGateR: semi-automated marker gating for single-cell multiplexed imaging. *Bioinformatics*, *40*(6). https://doi.org/10.1093/bioinformatics/btae356
